---
title: "Simple Parametric Tests"
# subtitle: "And Introduction to the Course and The Importance of Proper Statistics"
author: "Erik Kusch"
date: ""
fontsize: 10pt
output: 
  beamer_presentation: # handouts can be established by setting classoption: "handout" instead of classoption: t
    keep_tex: true
    toc: false # this is added through a later command
    slide_level: 3 # at how many pound signs (#) to assume slide title level
    includes:
      in_header: Style.tex
classoption: c
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=TRUE, tidy.opts=list(width.cutoff=50),tidy=TRUE, fig.height=8)
options(digits=2, width = 60)

# function to load packages and install them if they haven't been installed yet
install.load.package <- function(x) {
  if (!require(x, character.only = TRUE))
    install.packages(x)
  require(x, character.only = TRUE)
}
package_vec <- c("nonpar" # needed for Cochran's Q
                 )
sapply(package_vec, install.load.package)
```

<!--- ####### TOC (use this if you want ## headers to be included in the toc)--------------------
---------------------------------------  --->

\tableofcontents[subsubsectionstyle=hide]

<!--- ####### Inferential Statistics-------------------------------------------------------------
---------------------------------------  --->
# Background
### Introduction
\begin{center}
Parametric test are those statistical approaches which rely on \textbf{assumptions} about the parameters which define a population.
\end{center}
\vspace{.3cm}
\pause
Prominent parametric tests include:
\begin{itemize}
\item Pearson correlation (Seminar 9 - Correlation Tests)
\item \textbf{t-Test}
\item \textbf{An}alysis \textbf{O}f \textbf{Va}riance (ANOVA)
\item Linear regression
\item Multivariate extensions of parametric methods
\item ...
\end{itemize}

### Terminology
A reminder about the distinction of parametric and non-parametric tests (taken from Seminar 6):
\pause
\vspace{0.4cm}
\begincols[T]
  \begincol{.5\linewidth}
  \pause
  \textbf{Non-Parametric Tests}
    \begin{itemize}
    \item Less \textit{restrictive}
    \item Make \textit{little to no assumptions}
    \item Often a \textbf{black box}
    \item Require \textit{more data}
    \end{itemize}
  \endcol
  \begincol{.5\linewidth}
  \pause
  \textbf{Parametric Tests}
    \begin{itemize}
    \item More \textit{restrictive}
    \item Make \textit{strict assumptions}
    \item \textbf{Easy to interpret}
    \item Require \textit{less data}
    \end{itemize}
  \endcol
\endcols
\vspace{0.8cm}
\pause
$\rightarrow$ Parametric tests are \textbf{numerous}!

# Analyses
## t-Test (unpaired)
### Purpose And Assumptions
\textbf{t-Test} (unpaired) \flushright `t.test(..., paired = FALSE)` in base `R`  
\vspace{-10pt}
\small
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} &  To identify whether groups of variable values are different from one another. \vspace{5pt} \\
    \pause
    $H_0$ & \textit{There is no difference in characteristics of the response variable values in dependence of the classes of the predictor variable.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Predictor variable is binary
      \item Response variable is metric and \textbf{normal distributed} within their groups
      \item Variable values are \textbf{independent} (not paired)
      \end{itemize}
     \\
  \end{tabular}}
\end{table}
\pause
\scriptsize
$\rightarrow$ Test whether variance of response variable values in groups are equal (`var.test()`) and adjust `t.test()` argument `var.equal` accordingly.

### Minimal Working Example
\small
Let's feed data to our `t.test(..., paired = FALSE)` function that holds two groups with clearly differing means:
\vspace{.1cm}
\scriptsize
```{r tTest1}
data <- c(
  rnorm(10,5,1),
  rnorm(10,10,1)
)
factors <- as.factor(rep(c("A", "B"), each = 10))
t.test(data ~ factors, paired = FALSE)
```
\small
\vspace{.1cm}
The output above tells us that the means of our two groups are significantly different.


## t-Test (paired)
### Purpose And Assumptions
\textbf{t-Test} (paired) \flushright `t.test(..., paired = TRUE)` in base `R`  
\vspace{-10pt}
\small
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} &  To identify whether groups of variable values are different from one another. \vspace{5pt} \\
    \pause
    $H_0$ & \textit{There is no difference in characteristics of the response variable values in dependence of the classes of the predictor variable.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Predictor variable is binary
      \item Response variable is metric 
      \item \textit{Difference of response variable pairs} is \textbf{normal distributed}
      \item Variable values are \textbf{dependent} (paired)
      \end{itemize}
     \\
  \end{tabular}}
\end{table}
\pause
\scriptsize
$\rightarrow$ Test whether variance of response variable values in groups are equal (`var.test()`) and adjust `t.test()` argument `var.equal` accordingly.

### Minimal Working Example
\small
Let's feed data to our `t.test(..., paired = TRUE)` function that holds two connected groups with clearly differing means:
\vspace{.1cm}
\scriptsize
```{r tTest2}
data <- c(
  rnorm(10,5,1),
  rnorm(10,10,1)
)
factors <- as.factor(rep(c("A", "B"), each = 10))
t.test(data ~ factors, paired = TRUE)
```
\small
\vspace{.1cm}
The output above tells us that the means of our two connected groups are significantly different.

## Analysis of Variance (ANOVA)
### Introduction to ANOVA
\begin{center}
ANOVAs are used to test whether there is a difference between groups of variable values.
\end{center}
\vspace{.3cm}
\pause
There are \textbf{multiple versions of ANOVAs}:
\begin{itemize}
\item One-way ANOVA (one predictor variable)
\item Two-Way ANOVA (multiple predictor variables)
\item MANOVA (multivariate ANOVA/multiple response variables)
\item ANCOVA (categorical and continuous predictor variables)
\item MANCOVA (multivariate ANCOVA)
\end{itemize}

### Data for ANOVA
We will use the `crabs` data set from the `MASS` package

```{r ANOVACrabs}
library(MASS)
data(crabs)
head(crabs)
```

## One-Way ANOVA
### Purpose And Assumptions
\textbf{One-Way ANOVA} \flushright `anova()` in base `R`  
\vspace{-10pt}
\small
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} & To explain the variance of a continuous response variable in relation to one predictor variables. \vspace{5pt} \\
    \pause
    $H_0$ & \textit{Variance of response variable values is equal between levels of predictor variable.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Predictor variable is categorical
      \item Response variable is metric
      \item \textit{Response variable residuals} are \textbf{normal distributed}
      \item Variance of populations/samples are equal (\textbf{homogeneity})
      \item Variable values are \textbf{independent} (not paired)
      \end{itemize}
     \\
  \end{tabular}}
\end{table}
\pause
\scriptsize
$\rightarrow$ Test whether residuals are normal distributed with `shapiro.test()` in base `R`, test for homogeneity with `leveneTest()` in the `car` package.

### Minimal Working Example - Assumptions
Let's test whether body depth (BD) of crabs are varying when grouped by sex:
\vspace{.1cm}
\tiny
```{r One-Way}
OneWay <- with(crabs,  lm(BD ~ sex)) # MODEL
```
\tiny
\begincols[T]
  \begincol{.5\linewidth}
  \tiny
```{r One-Waya, tidy = FALSE, fig.height = 5.5}
plot(OneWay, 2)# Normality
shapiro.test(residuals(OneWay))
```
\endcol
\begincol{.5\linewidth}
```{r One-Wayb, tidy = FALSE, fig.height = 5.5}
plot(OneWay, 3)# Homogeneity
library("car")
leveneTest(BD ~ sex, data = crabs)
```  
\endcol
\endcols

\normalsize
\vspace{.2cm}
All good on the assumption check!

### Minimal Working Example - Analysis
Now let's run the analysis:
```{r One-Wayc}
anova(OneWay)
```
\vspace{.1cm}
As we can see, sex does not make for a statistically significant predictor of crab body depth.

### Minimal Working Example - Interpretation
\scriptsize
Let's interpret the result anyways:
\tiny
```{r One-Wayd}
summary(OneWay)
```
\vspace{.1cm}
\scriptsize
- Female crabs are estimated to have a body depth of `r OneWay$coefficients[1]`cm (**Intercept**) with males being `r OneWay$coefficients[2]`cm bigger, on average.  
- While we can be certain of the female estimate, we cannot say the same about the different to males.

## Two-Way ANOVA
### Purpose And Assumptions
\textbf{Two-Way ANOVA} \flushright `anova()` in base `R`  
\vspace{-10pt}
\small
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} & To explain the variance of a continuous response variable in relation to multiple predictor variables. \vspace{5pt} \\
    \pause
    $H_0$ & \textit{Variance of response variable values is equal between levels of predictor variables.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Predictor variables are categorical
      \item Response variable is metric
      \item \textit{Response variable residuals} are \textbf{normal distributed}
      \item Variance of populations/samples are equal (\textbf{homogeneity})
      \item Variable values are \textbf{independent} (not paired)
      \end{itemize}
     \\
  \end{tabular}}
\end{table}
\pause
\scriptsize
$\rightarrow$ Test whether residuals are normal distributed with `shapiro.test()` in base `R`, test for homogeneity with `leveneTest()` in the `car` package.

### Minimal Working Example - Assumptions
Let's test whether body depth (BD) of crabs are varying when grouped by sex and species as well as their interaction:
\vspace{.1cm}
\tiny
```{r Two-Way}
TwoWay <- with(crabs,  lm(BD ~ sex*sp))
```
\tiny
\begincols[T]
  \begincol{.5\linewidth}
  \tiny
```{r Two-Waya, tidy = FALSE, fig.height = 5.5}
plot(TwoWay, 2)# Normality
shapiro.test(residuals(TwoWay))
```
\endcol
\begincol{.5\linewidth}
```{r Two-Wayb, tidy = FALSE, fig.height = 5.5}
plot(TwoWay, 3)# Homogeneity
library("car")
leveneTest(BD ~ sex*sp, data = crabs)
```  
\endcol
\endcols

\normalsize
\vspace{.2cm}
All good on the assumption check!

### Minimal Working Example - Analysis
Now let's run the analysis:
\scriptsize
```{r Two-Wayc}
anova(TwoWay)
```
\vspace{.1cm}
\normalsize
The output above tells us that species and the interaction effect of sex and species are meaningful for understanding body depth of crabs.

### Minimal Working Example - Interpretation
\tiny
```{r Two-Wayd}
summary(TwoWay)
```
\vspace{.1cm}
- Female crabs of species B are estimated to have a body depth of `r TwoWay$coefficients[1]`cm (**Intercept**) with males of species B being `r TwoWay$coefficients[2]`cm bigger, on average.  
- Female crabs of species O are estimated to have a body depth of `r TwoWay$coefficients[3]`cm bigger than their female species B counterparts.  
- The difference in sex- vs. species-dependant change in body depth is `r TwoWay$coefficients[4]`cm.
- All estimates are statistically significant.



## ANCOVA
### Purpose And Assumptions
\textbf{ANCOVA} \flushright `anova()` in base `R`  
\vspace{-10pt}
\small
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} & To explain the variance of a continuous response variable in relation to mixed (continuous and categorical) predictor variables. \vspace{2pt} \\
    \pause
    $H_0$ & \textit{Adjusted variance and means of response variable values is equal between levels of predictor variables.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Predictor variables are categorical or continuous
      \item Response variable is metric
      \item \textit{Response variable residuals} are \textbf{normal distributed}
      \item Variance of populations/samples are equal (\textbf{homogeneity})
      \item Variable values are \textbf{independent} (not paired)
      \item Relationship between the response and covariate is linear.
      \end{itemize}
     \\
  \end{tabular}}
\end{table}
\vspace{-.2cm}
\pause
\tiny
$\rightarrow$ Test whether residuals are normal distributed with `shapiro.test()` in base `R`, test for homogeneity with `leveneTest()` in the `car` package.

### Minimal Working Example - Assumptions
Let's test whether carapace length (CL) of crabs are varying when grouped by species and the carapace width as a covariate:
\vspace{.1cm}
\tiny
```{r Ancova}
Ancova <- with(crabs,  lm(CL ~ sp*CW))
```
\tiny
\begincols[T]
  \begincol{.5\linewidth}
  \tiny
```{r Ancovaa, tidy = FALSE, fig.height = 5.5}
plot(Ancova, 2)# Normality
shapiro.test(residuals(Ancova))
```
\endcol
\begincol{.5\linewidth}
```{r Ancovab, tidy = FALSE, fig.height = 5.5}
plot(Ancova, 3)# Homogeneity
library("car")
leveneTest(CL ~ sp, data = crabs)
```  
\endcol
\endcols

\normalsize
\vspace{.2cm}
Assumptions are met!

### Minimal Working Example - Analysis
Now let's run the analysis:
\scriptsize
```{r Ancovac}
anova(Ancova)
```
\vspace{.1cm}
\normalsize
The output above tells us that all of our model coefficients are significant.

### Minimal Working Example - Interpretation
\tiny
```{r Ancovad}
summary(Ancova)
```
\vspace{.1cm}
- Crabs of species B have an estimated carapace length of `r Ancova$coefficients[1]`cm when their carpace width would be 0cm (**Intercept**) with members of species B being `r Ancova$coefficients[2]`cm bigger, on average at 0cm carapace width.  
- For each additional cm in carapace width, carapacae length in species B increases by `r Ancova$coefficients[3]`cm.  
- For each additional cm in carapace width, carapacae length in species O increases by `r Ancova$coefficients[3]`cm more than in species B.  
- All estimates except for the species-difference are statistically significant.



# Our Data
## Choice Of Variables
### Variables We Can Use
\vspace{.7cm}
\begincols[T]
  \begincol{.5\linewidth}
    \textbf{Response variables} (metric)
    \visible<2->{
    \begin{itemize}
    \item Weight
    \item Height
    \item Wing Chord
    \item Nesting Height
    \item Number of Eggs
    \item Egg Weight
    \end{itemize} 
    }
\endcol
\begincol{.5\linewidth}
  \pause \pause
    \textbf{Predictor variables} (categorical)
    \visible<4->{
    \begin{itemize}
    \item Sex (binary)
    \item Climate (binary)
    \item Climate (3 levels - Continental, Semi-Coastal, Coastal)
    \item Home Range (3 levels - Small, Medium, Large)
    \item Site Index (11 levels)
    \item Predator Presence/Type (3 levels - Avian vs. Non-Avian vs. None)
    \end{itemize} 
    }
\endcol
\endcols

## Research Questions
### Research Questions And Hypotheses
\begin{center}
So which of our major research questions (seminar 6) can we answer?
\end{center}
\vspace{.2cm}
\scriptsize
\begincols[T]
  \begincol{.5\linewidth}
  \pause
    \textbf{unpaired t-Test}
    \begin{itemize}
    \item \textit{Climate Warming/Extremes:} Does sparrow morphology change depend on climate?
    \item \textit{Sexual Dimorphism:} Does sparrow morphology change depend on Sex?
    \end{itemize} 
    \tiny
    \raggedright Use the `1 - Sparrow_Data_READY.rds` data set for these analyses. \vspace{.4cm} \linebreak
    \scriptsize
  \pause
    \textbf{paired t-Test} (suppose a resettling program)
    \begin{itemize}
    \item \textit{Climate Warming/Extremes:} Does sparrow morphology change depend on climate?
    \end{itemize} 
    \tiny
    \raggedright Use the `2b - Sparrow_ResettledSIUK_READY.rds` data set for these analyses. \vspace{.4cm} \linebreak
    \scriptsize 
\endcol
\begincol{.5\linewidth}
\pause
    \textbf{One-Way ANOVA}
    \begin{itemize}
    \item \textit{Climate Warming/Extremes:} Does sparrow morphology depend on climate?
    \item \textit{Predation:} Does nesting height depend on predator characteristics?
    \end{itemize} 
  \pause
    \textbf{Two-Way ANOVA}
    \begin{itemize}
    \item \textit{Sexual Dimorphism:} Does sparrow morphology depend on population status and sex?
    \end{itemize}
  \pause
    \textbf{ANCOVA}
    \begin{itemize}
    \item \textit{Climate Warming/Extremes:} Do sparrow characteristics depend on climate and latitude?
    \end{itemize} 
  \tiny
    \raggedright Use the `1 - Sparrow_Data_READY.rds` data set for these analyses. \vspace{.4cm} \linebreak
    \scriptsize
\endcol
\endcols
\pause
\begin{tcolorbox}[colback=burgundy!5,colframe=burgundy!40!black,title= ]
\centering Remember to diligently \textbf{check assumptions}!
\end{tcolorbox}

