---
title: "" # needs to be empty since it is specified in Title.Rmd
author: ""
date: ""
documentclass: "book"
fontsize: 10pt
output: 
  pdf_document:
    includes:
      in_header: Style_Exercise.tex
subparagraph: true # this is needed as a workaround to get titlesec to work
geometry: "left= 2cm,right = 2cm, bottom = 1.5cm, top = 2cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=TRUE, tidy.opts=list(width.cutoff=90),tidy=TRUE, fig.height=4, fig.width = 12)
options(digits=2, width = 90)
```

<!--- ####### PREAMBLE -----------
--------------------------------------------->
\begin{center}
\includegraphics[width=0.42\linewidth]{Figures/UiA-logo.png}
\hspace{.05\linewidth}
\includegraphics[width=0.42\linewidth]{Figures/UL-logo.jpg}
\end{center}

\\

\vspace{5.5cm}

\begin{center}
\textsc{\LARGE{
\textbf{Data Visualisation}\\
\vspace{0.15cm}
Biostatistics for Beginners \\
\vspace{0.3cm}
- } \vspace{0.3cm}\\
\large{BASIC STATISTICS FOR BIOLOGISTS\\
}}
\end{center}

\vspace{6.5cm}

\raggedleft
\textbf{Erik Kusch}  
\textit{PhD Student}  
Aarhus University  
Department of Bioscience  
Section for Ecoinformatics \& Biodiversity  
Center for Biodiversity and Dynamics in a Changing World (BIOCHANGE)  
Ny Munkegade 116, Building 1540  
8000 Aarhus  
Denmark  
email: erik@i-solution.de  

\raggedright
\clearpage

<!--- ####### TOC, TOF, TOT, TOL ------------------------------------------------>  
\pagestyle{plain}
\pagenumbering{roman}

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\begin{LARGE}
 \textbf{Summary:}\\
 \vspace{0.2cm}
\end{LARGE}
These are the solutions to the exercises contained within the handout to “Data Visualisation” which walks you through the basics of data visualisation in `R`using `ggplot2`. The plots presented here are using data from the `iris` data set supplied through the `datasets` package. 

Keep in mind that there is probably a myriad of other ways to reach the same conclusions as presented in these solutions.


\setcounter{secnumdepth}{3} 
\setcounter{tocdepth}{4} 
\tableofcontents
\endgroup

<!--- ####### DOCUMENT ---------------------------------------------------------->
\pagestyle{fancy}
\pagenumbering{arabic}

# Packages
Recall the exercise that went along with the last seminar (Descriptive Statistics) where we learnt the difference between a basic and andvanced preamble for package loading in `R`. Here (and in future exercises) I will only supply you with the advanced version of the preamble.  

Now let's load the `ggplot2` package into our `R` session so we'll be able to use its functionality for data visualisation as well as the `datasets` package to get the `iris` dat set.

```{r Packages}
# function to load packages and install them if they haven't been installed yet
install.load.package <- function(x) {
  if (!require(x, character.only = TRUE))
    install.packages(x)
  require(x, character.only = TRUE)
}
# packages to load/install if necessary
package_vec <- c("ggplot2", "datasets")
# applying function install.load.package to all packages specified in package_vec
sapply(package_vec, install.load.package)
```

\let\cleardoublepage\relax

# Loading `R`-internal data sets (`iris`)
The `iris` data set is included in the `datasets` package in `R`. An `R`-internal data set is loaded through the command `data()`. Take note that you do not have to assign this command's output to a new object (via `<-`). Instead, the dataset is loaded to your current environment by its name (iris, in this case). Keep in mind that this **can override** objects of the same name that are already present in your current session of `R`.
```{r RLoad}
data("iris")
```

\let\cleardoublepage\relax

# Inspect the data set  
Since we know that `iris` is a dataset, we can be reasonably sure that this object will be complex enough to warrant using the `str()` function for inspection:
```{r RInspect}
str(iris)
```

The `iris` dataset contains four measurements (`Sepal.Length`, `Sepal.Width`, `Petal.Length`, `Petal.Width`) for 150 flowers representing three species of iris (*Iris setosa*, *versicolor* and *virginica*).

\newpage

# **Boxplot** of `Petal.Length` by `Species`  
```{r PlottingExercise1, fig.height=8, tidy=FALSE}
ggplot(iris, # the data set
       aes(x=Species, y=Petal.Length) # aesthetics
       ) + geom_boxplot() + # this is the end of the bare minimum plot
  theme_bw() + labs(title="Petal Length of three different species of Iris")
```

THis boxplot shows us exactly how the distributions of petal length measurements of our three species of Iris are differing from one another. Despite the obvious trend in the data, **be sure not to report results through figures alone!** We will find out how to test whether the pattern we can observe here holds up to scrutiny at a later point in time of our seminars.

\newpage

# **Scatterplot** of `Petal.Length` and `Petal.Width`  
```{r PlottingExercise2, tidy=FALSE}
ggplot(iris, # the data set
       aes(x=Petal.Width, y=Petal.Length) # aesthetics
       ) + geom_point() + # this is the end of the bare minimum plot
  theme_bw() + labs(title="Petal Width and Petal Length of three different species of Iris")
```


# **Scatterplot** of `Petal.Length` and `Petal.Width` grouped by `Species`  
```{r PlottingExercise3, tidy=FALSE}
ggplot(iris, # the data set
       aes(x=Petal.Width, y=Petal.Length, colour = Species) # aesthetics
       ) + geom_point() + # this is the end of the bare minimum plot
  theme_bw() + labs(title="Petal Width and Petal Length of three different species of Iris") + 
  theme(legend.justification=c(1,0), legend.position=c(1,0)) + # legend inside
  scale_color_discrete(name="Iris Species")  # Change legend title
```

\newpage

# Relationship of `Sepal.Length` and `Sepal.Width`  
```{r PlottingExercise4, tidy=FALSE}
ggplot(iris, # the data set
       aes(x=Sepal.Width, y=Sepal.Length) # aesthetics
       ) + geom_point() + geom_smooth() + # this is the end of the bare minimum plot
  theme_bw() + labs(title="Petal Width and Petal Length of three different species of Iris")
```

# Relationship of `Sepal.Length` and `Sepal.Width` (grouped by `Species`)
```{r PlottingExercise5, tidy=FALSE}
ggplot(iris, # the data set
       aes(x=Sepal.Width, y=Sepal.Length, colour = Species) # aesthetics
       ) + geom_point() + geom_smooth() + # this is the end of the bare minimum plot
  theme_bw() + labs(title="Petal Width and Petal Length of three different species of Iris") + 
  theme(legend.justification=c(1,0), legend.position=c(1,0)) + # legend inside
  scale_color_discrete(name="Iris Species")  # Change legend title
```