---
title: "Metric Tests (Two-Sample Situations)"
# subtitle: "And Introduction to the Course and The Importance of Proper Statistics"
author: "Erik Kusch"
date: "??/??/20??"
fontsize: 10pt
output: 
  beamer_presentation: # handouts can be established by setting classoption: "handout" instead of classoption: t
    keep_tex: false
    toc: false # this is added through a later command
    slide_level: 3 # at how many pound signs (#) to assume slide title level
    includes:
      in_header: Style.tex
classoption: t
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=TRUE, tidy.opts=list(width.cutoff=50),tidy=TRUE, fig.height=8)
options(digits=2, width = 60)

# function to load packages and install them if they haven't been installed yet
install.load.package <- function(x) {
  if (!require(x, character.only = TRUE))
    install.packages(x)
  require(x, character.only = TRUE)
}
package_vec <- c("nonpar" # needed for Cochran's Q
                 )
sapply(package_vec, install.load.package)
```

<!--- ####### TOC (use this if you want ## headers to be included in the toc)--------------------
---------------------------------------  --->

\tableofcontents[subsubsectionstyle=hide]

<!--- ####### Inferential Statistics-------------------------------------------------------------
---------------------------------------  --->
# Background
### Introduction
\begin{center}
Metric tests are used to compare parameters of metric/ordinal variable values among groups/individuals.
\end{center}
\pause
Prominent metric tests for two-sample situations include:
\begin{itemize}
\item \textbf{Mann-Whitney U Test}
\item \textbf{Wilcoxon Signed Rank Test}
\item t Test (dealt with in seminar 12)
\item ...
\end{itemize}

\pause

Some of these tests rely on the \textbf{assumption of independence}:

\begin{tcolorbox}[colback=burgundy!5,colframe=burgundy!40!black,title=]
\centering The assumption of independence is a \textbf{crucial prerequisite} to many statistical procedures!
\end{tcolorbox}

### Independence
\small
\begincols[T]
  \begincol{.5\linewidth}
  \pause
    \textbf{Theory:}
    \begin{itemize}
    \pause
    \item Even the smallest dependence in your data can turn into heavily biased results (which may be undetectable).
    \pause
    \item A dependence is a connection between/within the data.
    \pause
    \item The assumption of independence relies on the absence of any connection in your data that haven't accounted for in your approach (accounting for it is difficult).
    \end{itemize} 
\endcol
\begincol{.5\linewidth}
  \pause
    \textbf{Independent data:}
    \begin{itemize}
    \pause
    \item \textit{Between Groups}\\
    Groups of data records should be pulled from different individuals.
    \pause
    \item \textit{Within Groups}\\
    Data values within the same group are not to influence one another.
    \pause
    \item  \textit{Within Individuals}\\
    Data values recorded for one individual should not influence each other. This is often an issue with repeated measurement approaches.
    \end{itemize} 
\endcol
\endcols
\vspace{10pt}
\pause
\centering $\rightarrow$ Fixing this \textit{after data collection} is \textbf{almost impossible}!

# Analyses
## Mann-Whitney U Test / Wilcoxon Rank-Sum Test
### Purpose And Assumptions
\textbf{Mann-Whitney U Test} \flushright `wilcox.test(..., paired = FALSE)` in base `R`  
\vspace{-10pt}
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} &  To identify whether groups of variable values are different from one another. \vspace{5pt} \\
    \pause
    $H_0$ & \textit{There is no difference in characteristics of the response variable values in dependence of the classes of the predictor variable.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Predictor variable is binary
      \item Response variable is ordinal or metric
      \item Variable values are \textbf{independent} (not paired)
      \end{itemize}
     \\
  \end{tabular}}
\end{table}

### Minimal Working Example
Let's use the `wilcox.test(..., paired = FALSE)` function to test whether the medians of an unnamed variable of two unconnected populations (`a` and `b`) with 10 individuals each are truly different:
\scriptsize
```{r MannWhitney}
set.seed(42)
a <- rnorm(n = 10, mean = 10, sd = 3)
b <- rnorm(n = 10, mean = 5, sd = 3)
wilcox.test(a, b, paired = FALSE)
```

```{r p.valMann, echo=FALSE}
p.val <- wilcox.test(a,b, paired=FALSE)$p.value
```
\normalsize
The medians are significantly different (p = $`r p.val`$ ). Keep in mind that the \textbf{populations do not have to be of the same size} for this!

## Wilcoxon Signed Rank Test
### Purpose And Assumptions
\textbf{Wilcoxon Signed Rank Test} \flushright `wilcox.test(..., paired = TRUE)` in base `R`  
\vspace{-10pt}
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} &  To identify whether groups of variable values in a repeated sampling set-up are different from one another. \vspace{5pt} \\
    \pause
    $H_0$ & \textit{There is no difference in characteristics of the response variable values in dependence of the classes of the predictor variable.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Predictor variable is binary
      \item Response variable is ordinal or metric
      \item Variable values are \textbf{dependent} (paired)
      \end{itemize}
     \\
  \end{tabular}}
\end{table}

### Minimal Working Example
Let's use the `wilcox.test(..., paired = FALSE)` function to test whether the medians of an unnamed variable of two connected samples (`a` and `b`) with 10 individuals each (i.e. one re-sampled population) are truly different:
\scriptsize
```{r Wilcox}
set.seed(42)
a <- rnorm(n = 10, mean = 10, sd = 3)
b <- rnorm(n = 10, mean = 5, sd = 3)
wilcox.test(a, b, paired = TRUE)
```

```{r p.valWilcox, echo=FALSE}
p.val <- wilcox.test(a,b, paired=TRUE)$p.value
```
\normalsize
The medians are significantly different (p = $`r p.val`$ ). Keep in mind that the \textbf{samples have to be of the same size} for this (i.e. there is one data record in `b` that corresponds to one data record in `a`)!

# Our Data
## Choice Of Variables
### Variables We Can Use

\begincols[T]
  \begincol{.5\linewidth}
    \textbf{Response variables} (metric/ordinal)
    \visible<2->{
    \begin{itemize}
    \item Weight
    \item Height
    \item Wing Chord
    \item Nesting Height
    \item Number of Eggs
    \item Egg Weight
    \item Home Range
    \end{itemize} 
    }
\endcol
\begincol{.5\linewidth}
  \pause \pause
    \textbf{Predictor variables} (binary)
    \visible<4->{
    \begin{itemize}
    \item Population Status (Introduced vs. Native)
    \item Sex (Male vs. Female)
    \item Nesting Site (Tree vs. Shrub)
    \item Predator Presence (Yes vs. No)
    \item Predator Type (Avian vs. Non-Avian)
    \item Climate (Continental vs. Coastal)
    \end{itemize} 
    }
\endcol
\endcols

## Research Questions
### Research Questions And Hypotheses
\begin{center}
So which of our major research questions (seminar 6) can we answer?
\end{center}

\small
\begincols[T]
  \begincol{.5\linewidth}
  \pause
    \textbf{Mann Whitney U Test}
    \begin{itemize}
    \item \textit{Climate Warming/Extremes:} Does sparrow morphology depend on climate?
    \item \textit{Predation:} Does nesting height depend on predator characteristics?
    \item \textit{Competition:} Does home range depend on climate?
    \item \textit{Sexual Dimorphism:} Does sparrow morphology depend on sex?
    \end{itemize} 
\endcol
\begincol{.5\linewidth}
  \pause
    \textbf{Wilcoxon Signed Rank Test} (suppose a resettling program)
    \begin{itemize}
    \item \textit{Climate Warming/Extremes:} Does sparrow morphology change depend on climate?
    \item \textit{Predation:} Does nesting height depend on predator characteristics?
    \item \textit{Competition:} Does home range depend on climate?
    \end{itemize} 
    \footnotesize
$\rightarrow$ Find the data in \linebreak \textit{2 - Sparrow\_Resettled\_READY.rds}
\endcol
\endcols

# Exercise
### The `with()` function
The `with()` function can be used to **make** your **code**:  
- easier to **write** and **read**  
- more **accessible**  

\vspace{10pt}

\pause

You might hear someone refer to it as *soft attach* because it **works** a lot **like** the `attach()` function in `R` but **causes none of its problems**

\vspace{10pt}

\pause

You use `with()` to refer to data contained within a data object inside `R`:
```{r with, eval=FALSE, tidy=FALSE}
with(data.object, 
     expression(reference.to.object.within.data.object)
     )
```