---
title: "Descriptive Statistics"
# subtitle: "An Introduction to the Course and The Importance of Proper Statistics"
author: "Erik Kusch"
date: ""
fontsize: 10pt
output: 
  beamer_presentation: # handouts can be established by setting classoption: "handout"
    keep_tex: false
    toc: false # this is added through a later command
    slide_level: 3 # at how many pound signs (#) to assume slide title level
    includes:
      in_header: Style.tex
classoption: t
---

```{r setup, include=FALSE}
options(scipen=1, digits=4)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=TRUE, tidy.opts=list(width.cutoff=20),tidy=TRUE, fig.height=8, digits=4)
```

<!--- ####### TOC (use this if you want ## headers to be included in the toc)--------------------
---------------------------------------  --->

\tableofcontents[subsubsectionstyle=hide]

<!--- ####### Descriptive Statistics-------------------------------------------------------------
---------------------------------------  --->
# Background
## Introduction
### Introduction
\begin{tcolorbox}[colback=bondiblue!5,colframe=bondiblue!40!black,title=]
\centering Descriptive statistics are used to \textbf{summarize data}.
\end{tcolorbox}

\vspace{-0.3cm}
\pause 

\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \textit{The aim:} &  To describe a given set of data records $n$ in regard to a certain \textbf{variable} $p_j$ or set of variables $p$.\\
           \pause
             & \\
           \textit{The procedure:} & Using an adequately chosen set of methods to \textbf{summarize or visualize} the data at hand. \\
          \end{tabular}}
        \label{Des_Stats}
\end{table}

\vspace{0.2cm}

\pause
\begin{tcolorbox}[colback=bondiblue!5,colframe=bondiblue!40!black,title=]
\centering Characteristics of \textbf{variables} are often expressed via \textbf{parameters}.
\end{tcolorbox}

## Methods Of Descriptive Statistics
### Methods \& Quirks
Information is usually handed to descriptive statistics as $n \times p$ (row count $\times$ column count) data frames.
\pause
\hfill \linebreak
This information is used to calculate informative \textbf{parameters}:

\begin{columns}[T]
  \begin{column}{.5\linewidth}
  \visible<3->{\textbf{Location Parameters} (Measures Of Central Tendency):}
    \begin{itemize}
    \visible<4->{
    \item Arithmetic Mean
    \item Mode
    \item Median
    \item Minimum, Maximum, Range
    \item ...
    }
    \end{itemize}
  \end{column}
  \begin{column}{.5\linewidth}
  \visible<3->{\textbf{Dispersion Parameters} (Measures Of Spread): }
    \begin{itemize}
    \visible<5->{
    \item Variance
    \item Standard Deviation
    \item Quantile Range
    \item ...
    }
    \end{itemize}
  \end{column}
\end{columns}

\vspace{0.2cm}
\pause \pause \pause \pause
\begin{tcolorbox}[colback=burgundy!5,colframe=burgundy!40!black,title=]
\centering Descriptive statistics \textbf{do not allow generalisation} beyond the data!
\end{tcolorbox}

# Parameters In Descriptive Statistics
## What Are Parameters?
### Parameters And Their Meaning
\textbf{What is a parameter?}    
In the case of descriptive statistics, a \textit{parameter} presents some information on the shape of the distribution of the values of a certain variable.

\vspace{0.4cm}
\pause

\textbf{What's the fuss?}  
Parameters can be used to summarise data properties and make large data sets with a multitude of values per variable more accessible. 

\vspace{0.4cm}
\pause

\textbf{So?}  
To know which parameters to use one must know which ones there are and how to calculate them.

\vspace{0.4cm}
\pause
\begin{tcolorbox}[colback=bondiblue!5,colframe=bondiblue!40!black,title=]
\centering Parameters are, more or less, \textbf{digested data}.
\end{tcolorbox}

## Getting Practical
### Creating Some Data
For the following computation of descriptive statistics parameters, we will need the following data:
\footnotesize
```{r Data, tidy.opts=list(width.cutoff=55)}
set.seed(42) # making the code reproducible
data_vec <- rnorm(mean = 20, sd = 2, n = 54)
matrix(sort(data_vec), nrow = 6)
```
\normalsize
\vspace{.5cm}
$\rightarrow$ Calculation of parameters of descriptive statistics is reserved almost exclusively for `numeric` data records

## Location Parameters/Measures Of Central Tendency
### Arithmetic Mean (Theory)
\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \textit{Definition:} & Also called average, this metric is the mathematical average of the given data values. \\
           & \textbf{Non-resistant to outliers and asymmetric distributions.}\\
             & \\
           \hline
             & \\
           \textbf{\textit{Calculation:}} & $\overline{x} = \mu = \frac{1}{n}\sum\limits_{i=1}^nx_i$ \\
             & \\
           $\overline{x} = \mu$  & Arithmetic mean \\
           $n$  & Number of samples ($=$ number of values for the variable in question) \\
           $i$  & Index of variable values ($i = 1,2, .., n$) \\
           $x_i$  & $i$\textsuperscript{th} value of variable $x$ \\
          \end{tabular}}
        \label{Des_Stats}
\end{table}

### Arithmetic Mean (Calculation in `R`)
The arithmetic mean is calculated using the `mean()` function contained within base `R`.
\begincols[T]
  \begincol{.5\linewidth}
```{r mean, eval=TRUE}
# calculation
mean(data_vec)
```
  \endcol
  \begincol{.5\linewidth}
```{r meanplot, echo=FALSE}
plot(density(data_vec), main = "Mean of data_vec")
abline(v = mean(data_vec), col = "red")
```
  \endcol
\endcols

### Median (Theory)
\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \textit{Definition:} &  The median is the value separating the higher half of the data values from the lower half.\\
           & \textbf{Resistant to outliers and asymmetric distributions.}\\
             & \\
           \hline
             & \\
           \textbf{\textit{Calculation:}} & $median(x) = (\frac{n+1}{2})^{th} \quad \text{odd numbers of data values}$ \\
            & $median(x) = \frac{\big((\frac{n}{2})^{th} + (\frac{n}{2}+1)^{th} \big)}{2} \quad \text{even numbers of data values}$ \\
             & \\
           $median(x)$  & Median of the values available for variable $x$ \\
           $n$  & Number of observations available for $x$ \\
          \end{tabular}}
        \label{Des_Stats}
\end{table}

### Median (Calculation in `R`)
The median is calculated using the `median()` function contained within base `R`.
\begincols[T]
  \begincol{.5\linewidth}
```{r median, eval=TRUE}
# calculation
median(data_vec)
```
  \endcol
  \begincol{.5\linewidth}
```{r medianplot, echo=FALSE}
plot(density(data_vec), main = "Median of data_vec")
abline(v = median(data_vec), col = "blue")
```
  \endcol
\endcols

### Mode (Theory)
\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \textit{Definition:} & The mode of a set of data values is the value that is the most common. \\
           & \textbf{Resistant to outliers but the shape of the distribution might be crucial.}\\
             & \\
           \hline
             & \\
           \textbf{\textit{Calculation:}} & $mode(x) = max_{k=1}\big(I_{i=1}(x_i = x_k)\big)$ \\
             & \\
           $mode(x)$  & Mode of the values available for variable $x$ \\
           $max_{k=1}()$  & Maximising argument for $k$ in 1 to $p$ \\
           $I_{i}()$ & Identifier that returns 1 if the internal statement is true with $i$ in 1 to $n$  \\
          \end{tabular}}
        \label{Des_Stats}
\end{table}

### Mode (Calculation in `R`)
One may wish to use the `max()` and `table()` function contained within the base `R` or through `mlv(..., method="mfv")` within the `modeest` package:
\footnotesize
\begincols[T]
  \begincol{.5\linewidth}
```{r mode, eval=TRUE, tidy = FALSE}
# counts of values in rounded vector
table <- table(round(data_vec))
table # counts
# most common appearance
max <- max(table)
max # maximum appearances
```
  \endcol
  \begincol{.5\linewidth}
```{r mode2, eval=TRUE, tidy = FALSE}
# position of maximum in table
pos <- which(table == max) 
pos # mode position
# value at maximum position
mode <- names(table)[pos]
as.numeric(mode) # mode
```
  \endcol
\endcols

### Minimum, Maximum, Range (Theory)
Sometimes, one may want to use the following, simple information on data values:
\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \textit{Maximum:} &  The highest value available for a given variable.\\
             & \\
           \textit{Minimum:} &  The lowest value available for a given variable.\\
             & \\
           \textit{Range:} &  The span of values that the data distribution defined by minimum and maximum extends over.\\
             & \\
          \end{tabular}}
        \label{Des_Stats}
\end{table}


### Minimum, Maximum, Range (Calculation in `R`)
\begincols[T]
  \begincol{.5\linewidth}
```{r minmaxrange, eval=TRUE}
# calculation
min(data_vec)
max(data_vec)
range(data_vec)
```
  \endcol
  \begincol{.5\linewidth}
```{r minmaxrangeplot, echo=FALSE}
par(mfrow = c(1, 1))
plot(density(data_vec), main = "Minimum and Maximum of data_vec")
abline(v = max(data_vec), col = "orange")
abline(v = min(data_vec), col = "orange")
```
  \endcol
\endcols

### Which Location Parameter Do I Use?
All measures of central tendency describe the central position of a frequency distribution of values of a given variable in the data set at hand.

\vspace{0.7cm}
\pause

The \textit{arithmetic mean} is only really useful when concerned with symmetric distributions of data values.  

\vspace{0.5cm}
\pause

The \textit{median} exhibits robust behaviour when faced with asymmetric distributions of data values.

\vspace{0.5cm}
\pause

The \textit{mode} is most applicable to the classification setting and rarely used.

\vspace{0.7cm}
\pause

\centering $\rightarrow$ The \textbf{median} will usually do.

### You Can Do It Yourself

\centering \textbf{Remember:} You can code most of these basic parameter calculations yourself.
\visible<2->{
\centering \includegraphics[width=0.6\linewidth]{Figures/CodeHacker.jpg}}

## Dispersion Parameters/Measures Of Spread
### Variance (Theory)
\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \textit{Definition:} & Variance measures how much data values are spread out from their average value. \\
             & \textbf{Non-resistant to outliers and asymmetric distributions.}\\
             & \\
           \hline
             & \\
           \textbf{\textit{Calculation:}} & $s^2 = \frac{1}{n-1}\sum\limits_{i=1}^n(x_i - \overline{x})^2$ \\
             & \\
           $s^2$  & Variance \\
           $n$  & Number of samples ($=$ number of values for the variable in question) \\
           $i$  & Index of variable values ($i = 1,2, .., n$) \\
           $x_i$  & $i$\textsuperscript{th} value of variable $x$ \\
           $\overline{x}$  & Arithmetic mean \\
          \end{tabular}}
        \label{Des_Stats}
\end{table}

### Variance (Calculation in `R`)
The variance is calculated using the `var()` function contained within base `R`.
\begincols[T]
  \begincol{.5\linewidth}
```{r variance, eval=TRUE}
# calculation
var(data_vec)
```
  \endcol
  \begincol{.5\linewidth}
```{r varianceplot, echo=FALSE}
par(mfrow = c(1, 1))
plot(density(data_vec), main = "Variance of data_vec")
abline(v = mean(data_vec), col = "red")
abline(v = mean(data_vec)-var(data_vec)/2, col = "purple")
abline(v = mean(data_vec)+var(data_vec)/2, col = "purple")
```
  \endcol
\endcols
\centering Note that his plot shows the span of the variation around the mean.

### Standard Deviation (Theory)
\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \textit{Definition:} & The standard deviation quantifies the amount of variation or dispersion of a set of data values.\\
             & \textbf{Non-resistant to outliers and asymmetric distributions.}\\
             & \\
           \hline
             & \\
           \textbf{\textit{Calculation:}} & $SD = s = \sqrt{s^2}$\\
             & \\
           $SD = s$ & Standard Deviation\\
           $s^2$  & Variance \\
          \end{tabular}}
        \label{Des_Stats}
\end{table}

### Standard Deviation (Calculation in `R`)
The standard deviation is calculated using the `sd()` function contained within base `R`.
\begincols[T]
  \begincol{.5\linewidth}
```{r SD, eval=TRUE}
# calculation
sd(data_vec)
```
  \endcol
  \begincol{.5\linewidth}
```{r SDplot, echo=FALSE}
par(mfrow = c(1, 1))
plot(density(data_vec), main = "Standard Deviation of data_vec")
abline(v = mean(data_vec), col = "red")
abline(v = mean(data_vec)-sd(data_vec), col = "purple")
abline(v = mean(data_vec)+sd(data_vec), col = "purple")
```
  \endcol
\endcols
\centering Note that his plot shows the span of one standard deviation above and below the mean.

### Quantile Range  (Theory)
\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \textit{Definition:} & Quantiles are cut points dividing the range of a distribution of data values into adjacent intervals with equal probabilities. You will always receive one cut-point less than quantiles are produced.\\
             & \textbf{Resistant to outliers and asymmetric distributions.}\\
             & \\
           \hline
          \end{tabular}}
        \label{Des_Stats}
\end{table}
\pause
Most often, one uses the following quantiles:
\begin{itemize}
\item \textit{Quantile 50:} This is basically the median.
\item \textit{Quantile 25 and 75:} These are also known as \textit{quartiles}.
\end{itemize}


### Quantile Range  (Calculation in `R`)
Quantiles are calculated using the `quantile()` function contained within base `R`. A second argument, within this function can be specified to call certain quantiles.
\begincols[T]
  \begincol{.5\linewidth}
```{r quant, eval=TRUE, tidy=FALSE}
# quantiles we want
q <- c(0.25, 0.5, 0.95, 0.99)
# calculation
quantile(data_vec, q)
```
  \endcol
  \begincol{.5\linewidth}
```{r quantplot, echo=FALSE}
par(mfrow = c(1, 1))
plot(density(data_vec), main = "Quantiles of data_vec")
abline(v = mean(data_vec), col = "red")
abline(v = as.numeric(quantile(data_vec, c(0.25, 0.5, 0.95, 0.99))[1]), col = "purple")
abline(v = as.numeric(quantile(data_vec, c(0.25, 0.5, 0.95, 0.99))[2]), col = "purple")
abline(v = as.numeric(quantile(data_vec, c(0.25, 0.5, 0.95, 0.99))[3]), col = "purple")
abline(v = as.numeric(quantile(data_vec, c(0.25, 0.5, 0.95, 0.99))[4]), col = "purple")
```
  \endcol
\endcols

### Which Dispersion Parameter Do I Use?
All measures of spread describe the spread of a frequency distribution of values of a given variable in the data set at hand.

\vspace{0.7cm}
\pause

The \textit{variance} is only really useful when concerned with symmetric distributions of data values.  

\vspace{0.5cm}
\pause

The \textit{standard deviation} is only really useful when concerned with symmetric distributions of data values.  

\vspace{0.5cm}
\pause

The \textit{quantiles} exhibit robust behaviour when faced with asymmetric distributions of data values.

\vspace{0.7cm}
\pause

\centering $\rightarrow$ The \textbf{quantiles} will usually do.


## Making Life Easier in `R`
### The `summary()` Function
```{r summary, eval=TRUE}
# calculation
summary(data_vec)
```

\begincols[T]
  \begincol{.5\linewidth}
  \vspace{.5cm}
  The `summary()`function can be called on a vector object in `R` to return some of the most useful information on measures of central tendency and measures of spread.
  \endcol
  \begincol{.5\linewidth}
```{r summaryplot, echo=FALSE, fig.height=6}
par(mfrow = c(1, 1))
plot(density(data_vec), main = "Summary of data_vec")
abline(v = mean(data_vec), col = "red")
abline(v = as.numeric(summary(data_vec)[1]), col = "orange")
abline(v = as.numeric(summary(data_vec)[2]), col = "purple")
abline(v = as.numeric(summary(data_vec)[3]), col = "blue")
abline(v = as.numeric(summary(data_vec)[4]), col = "red")
abline(v = as.numeric(summary(data_vec)[5]), col = "purple")
abline(v = as.numeric(summary(data_vec)[6]), col = "orange")
```
  \endcol
\endcols

# Exercise
## Excel data
### Loading Excel data into `R`
Excel is a **valuable tool** for **data accquisition** but almost **useless** when it comes to **statistical analyses** or **data visualisation** in biological sciences.

\vspace{0.5cm}
\pause
\centering So **how do you get your excel data into `R`**?

\vspace{0.2cm}
\pause

\raggedright
Loading procedure depends on **file format**:
\begincols[T]
  \begincol{.5\linewidth}
  \pause
  **.csv** - I recommend using this format as it allows for less alteration and is compressed.  
  *Functions*: `read.csv()` and `read.table()` (also works on .txt files)  
  \endcol
  \begincol{.5\linewidth}
  \pause
  **.xls, .xlsx, etc.** - Go for this if you need to alter your data by hand (which you shouldn't. EVER!).  
  *Functions*: `read.xlsx()` (included in `xlsx` package)  
  \endcol
\endcols

\pause
\vspace{0.5cm}
\centering You can also **use `R`** to **save data in excel format**.

### Inspecting Data
The **most common** form of data is the **`data frame`** which you can:

\vspace{0.5cm}
\begincols[T]
  \begincol{.5\linewidth}
  \pause
  **Inspect** using functions such as:  
  - `dim()` to access the dimensions  
  - `str()` to access types and modes  
  - `colnames()`/`rownames()` to asses column and row names  
  - `head()`/`tail()` to show only the top or bottom five rows of the data set  
  - `table()` to show a count of items in a vector
  
  \endcol
  \begincol{.5\linewidth}
  \pause
  **Subset** using the different sub-setting methods:  
  - `[r,c]` can be used to index rows (r) and columns (c)  
  - `$` can be used to index column names  
  
  \endcol
\endcols

\pause
\vspace{.5cm}
\centering $\rightarrow$ This is also how you **extract data from a `data frame`** (and most objects within `R`).


## Parameters
### Calculating parameters of descriptive statistics

Your **ToDo-List** for this exercise:  

\vspace{.1cm}

* Load the file DescriptiveData.csv into `R`  

\vspace{.2cm}

* Identify what kind of information it contains  

\vspace{.2cm}

* Calculate the location parameters and parameters of spread for any of the variables contained within the data set that catch your interest.  

\vspace{.2cm}

* Question the validity of your findings and the data

\vspace{.5cm}
\pause
The solution file will deal with all of the variables contained within the data set so don't worry about which one to pick and just **have fun**.