---
title: "Introduction to R"
subtitle: 'First Steps in R'
author: "Erik Kusch"
date: ""
documentclass: "report"
fontsize: 10pt
output: 
  # rmdformats::material:
  html_document:
    self_contained: true
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = TRUE, fig.height = 4.5)
```

These are the solutions to the exercises contained within the handout to Introduction to R which walks you through the basics of the `R` machinery. `R` is a coding language that can be highly individualised and hence there are often multiple solutions to the same problem. Within these solutions, I shall only present you with one solution for every given task. However, do keep in mind that there is probably a myriad of other ways to achieve your goal.

# Creating and Inspecting Objects
## Vector
- A vector reading: "A", "B", "C"  
```{r createVectors1}
Letters_vec <- c("A", "B", "C")
Letters_vec
length(Letters_vec)
```
- A vector reading: 1, 2, 3  
```{r createVectors2}
Numbers_vec <- c(1, 2, 3)
Numbers_vec
length(Numbers_vec)
```
- A vector reading: `TRUE`, `FALSE`  
```{r createVectors3}
Logic_vec <- c(TRUE, FALSE)
Logic_vec
length(Logic_vec)
```
- A vector of the elements of the first three vectors
```{r createVectors4}
Big_vec <- c(Letters_vec, Numbers_vec, Logic_vec)
Big_vec
length(Big_vec)
```
- A vector reading as a sequence of full numbers from 1 to 20
```{r createVectors5}
Seq_vec <- c(1:20)
Seq_vec
length(Seq_vec)
```

## Factor
- A factor reading: "A", "B", "C"  
```{r createFactors1}
Letters_fac <- factor(x = c("A", "B", "C"))
Letters_fac
length(Letters_fac)
```
- A factor reading: 1, 2, 3  
```{r createFactors2}
Numbers_fac <- factor(x = c(1, 2, 3))
Numbers_fac
length(Numbers_fac)
```
- A factor reading: 1, 2, 3 but only levels 1 and 2 are allowed  
```{r createFactors5}
Constrained_fac <- factor(x = c(1,2,3), levels = c(1,2))
Constrained_fac
length(Constrained_fac)
```

- A factor reading: 1, 2, 3 levels 1 - 4 are allowed  
```{r createFactors6}
Expanded_fac <- factor(x = c(1,2,3), levels = c(1,2,3,4))
Expanded_fac
length(Expanded_fac)
```



## Matrix
- The first two vectors we established in distinct columns of a matrix
```{r createMatrices1}
Combine_mat <- matrix(data = c(Numbers_vec, Letters_vec), ncol = 2)
Combine_mat
dim(Combine_mat)
```
- The first two vectors we established in distinct rows of a matrix
```{r createMatrices2}
Pivot_mat <- matrix(data = c(Numbers_vec, Letters_vec), nrow = 2, byrow = TRUE)
Pivot_mat
dim(Pivot_mat)
```
- The above matrix with meaningful names
```{r createMatrices3}
Names_mat <- Pivot_mat
dimnames(Names_mat) <- list(c("Numbers", "Letters"))
Names_mat
dim(Names_mat)
```



## Data Frame
- The first matrix we established as a data frame
```{r createDFs1}
Combine_df <- data.frame(Combine_mat)
Combine_df
dim(Combine_df)
```
- The previous data frame with meaningful names
```{r createDFs2}
Names_df <- Combine_df
colnames(Names_df) <- c("Numbers", "Letters") 
Names_df
dim(Names_df)
```

## List
  - The first two vectors we created  
```{r createLists1}
Vectors_ls <- list(Numbers_vec, Letters_vec)
Vectors_ls
length(Vectors_ls)
```

# Statements and Loops
## Statements
- `Numbers_vec` contains more elements than `Letters_fac`?
```{r Sate1}
length(Numbers_vec) > length(Letters_fac)
```
- The first column of `Combine_df` is shorter than `Vectors_ls`?
```{r Sate2}
length(Combine_df[,1]) < length(Vectors_ls)
```
- The elements of `Letters_vec` are the same as the elements of `Letters_fac`?
```{r Sate3}
Letters_vec == Letters_fac
```

## Loops
- Print each element of `Vectors_ls`  
```{r Loop1}
for(i in 1:length(Vectors_ls)){
  print(Vectors_ls[[i]])
}
```
- Print each element of `Numbers_vec` + 1  
```{r Loop2}
Numbers_veca <- Numbers_vec +1
for(i in 1:length(Numbers_veca)){
  print(Numbers_veca[i])
}
```
- Subtract 1 from each element of the first column of `Combine_mat` and print each element separately
```{r Loop3}
Mat_column <- Combine_mat[,1] # extract data
Mat_column <- as.numeric(Mat_column) # convert to numeric
Mat_column <- Mat_column-1 # substract 1
for(i in 1:length(Mat_column)){
  print(Mat_column[i])
}
```
# Useful Commands
- Read out your current working directory (not showing you the result as it is different on every machine, it should start like this "C:/Users/....")
```{r Useful1, eval=FALSE}
getwd()
```
- Inspect the `Vectors_ls` object using the `View()` function (again, I am not showing you the result as this only works directly in `R` or Rstudio)
```{r Useful2, eval=FALSE}
View(Vectors_ls)
```
- Inspect the `Combine_df` object using the `View()` function (again, I am not showing you the result as this only works directly in `R` or Rstudio)
```{r Useful3, eval=FALSE}
View(Combine_df)
```
- Get the help documentation for the `as.matrix()` function (again, I am not showing you the result as this only works directly in `R` or Rstudio)
```{r Useful4, eval=FALSE}
?as.matrix
```
- Install and load the `dplyr` package  
```{r Useful5, eval=FALSE}
install.packages("dplyr")
library(dplyr)
```
- Remove the `Logic_vec` object from your working environment  
```{r Useful6}
rm(Logic_vec)
```
- Clear your entire working environment 
```{r Useful7}
ls() # this command shows you all the object in the environment
rm(list=ls())
```