---
title: "Introduction to R"
# subtitle: "And Introduction to the Course and The Importance of Proper Statistics"
author: "Erik Kusch"
date: ""
fontsize: 10pt
output: 
  beamer_presentation: # handouts can be established by setting classoption: "handout" instead of classoption: t
    keep_tex: true
    toc: false # this is added through a later command
    slide_level: 3 # at how many pound signs (#) to assume slide title level
    includes:
      in_header: Style.tex
classoption: "handout"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=TRUE, tidy.opts=list(width.cutoff=60),tidy=TRUE)
options(digits=2)
```

<!--- ####### TOC (use this if you want ## headers to be included in the toc)--------------------
---------------------------------------  --->

\tableofcontents[subsubsectionstyle=hide]

<!--- ####### Introduction to the Course---------------------------------------------------------
---------------------------------------  --->
# Objects in `R`
## The Basics
### Objects in `R`
`R` is an **object-based** programming language hence why *objects* are one of the *smallest* and *most important* units anyone attempting to handle `R` should be aware of. Like words in a sentence they **carry information**.  
\vspace{0.5cm} \pause
Objects are characterised by **attributes**:  
\vspace{-0.2cm}
\pause
\begin{small}
\begin{table}[h!]
  \begin{tabular}{p{2cm}p{6cm}p{3cm}}
    \textbf{Attribute} & Meaning & Definition \\\hline
    \textbf{Name} & R-internal name which is used to refer to the object in question & User-defined \\
    \textbf{Type} & R-internal class of the object in question & Often user-defined \\
    \textbf{Mode} & R-internal class of values contained within the object in question & Not user-defined \\
    \textbf{Dimensions} & Arrangement of content within the object in question & Often user-defined \\
  \end{tabular}
\end{table}
\end{small}

### Assigning and Removing Objects
An object is **assigned** some content in `R` as follows: 

*  Name `<-` Object/Content (this is the way to go)
*  Name `=` Object/Content (avoid this)  

\vspace{.4cm}

This will **add** an object of the chosen name **to** the **working environment**.  
\pause
\small
\begin{tcolorbox}[colback=burgundy!5,colframe=burgundy!40!black,title=]
Only one object of the same name can be present in the working environment at any given time.
\end{tcolorbox}

\vspace{.2cm}

\normalsize
\pause
To **alter/remove** a specific object from your `R` working environment do either of the following:

* Name `<-` Object/Content (simply **overwrite** it)
* `rm(`Name`)` (**remove** the object from your working environment)

## Basic `R`-internal Modes
### Object Modes
**Modes** refer to the class of object-contained values. Usually, you will only encounter some of the following basic modes:  

\pause
\small  

| `R`-internal object mode | Real-world counterpart |
|------:|:-----|
| `character` | A letter, word or sentence |
| `numeric` | A number (can have decimal points) |
| `logical` | An indicator of `TRUE` or `FALSE` |

\pause
\normalsize  

Within `R`, the **mode of any objects content** can be **identified** using the `class()` function after having broken down the object into basic `vector` type components.  
\pause
Additionally, one may want to employ the `str()` function which attempts to automatically perform some of the subsetting for you and give you an overview of the components within your object.

## Basic `R`-internal Types
### Object Types/Classes
Objects in `R` appear as different **types** (also referred to as *classes*):

\pause
\small  

| `R`-internal object type | Real-world counterpart |
|------:|:-----|
| `vector` | A list of variable values |
| `factor` | A list of variable values with pre-defined possible values |
| `matrix` | A table of variable values |
| `data frame` | A table of variable values |
| `function` | A recipe-style functional expression detailing a process |
| `list` | A list where each element corresponds to a list, table or recipe |

\normalsize
\pause

Within `R`, the **type of any object** can be **identified** using the `class()` or `str()` function. Note that, for an object of type `vector`, the `class()` function returns the **mode** of the vectors contents.  

### Vector I (`character`)
`Vectors` are created using the function `c(... , ...)` ("c" stands for *concatenate*; "," separates individual units within the vector) in `R`:

\vspace{.4cm}

* A `character` vector:  

\vspace{.2cm}

\scriptsize
```{r createVectors1}
Letters_vec <- c("A", "B", "C")
Letters_vec
class(Letters_vec)
```
\normalsize
\vspace{.4cm}
You can also **convert** object element mode to be of `character` by using the function `as.character()`.

### Vector II (`numeric`)
* A `numeric` vector:  

\vspace{.2cm}

\scriptsize
```{r createVectors2}
Numbers_vec <- c(1, 2, 3)
Numbers_vec
class(Numbers_vec)
```
\normalsize
\vspace{.4cm}
You can also **convert** object element mode to be of `numeric` by using the function `as.numeric()`.

### Vector III (`logical`)
* A `logical` vector:  

\vspace{.2cm}

\scriptsize
```{r createVectors3}
Logic_vec <- c(TRUE, FALSE)
Logic_vec
class(Logic_vec)
```
\normalsize
\vspace{.4cm}
You can also **convert** object element mode to be of `logical` by using the function `as.logical()` (keep in mind that this will only yield partially desirable results).

### Factor
`Factors` are created using the function `factor(x , levels , ...)` ("x" represents the data; "levels" indicates the preconceived levels our data should take and is usually estimated from the data by default) in `R`.  
\vspace{.4cm}

* A `factor` type object:  

\vspace{.2cm}
\scriptsize
```{r createFactors1, tidy=FALSE}
Letters_fac <- factor(x = c("A", "B", "C"))
Letters_fac
class(Letters_fac)
```
\normalsize
\vspace{.4cm}
You can also **convert** object types to be of `factor` by using the function `as.factor()`.

### Lists
`Lists` are created using the function `list(...)` ("..." represents the objects passed to the list) in `R`.
\vspace{.2cm}
\scriptsize
```{r createLists1}
Vectors_ls <- list(Numbers_vec, Letters_vec)
Vectors_ls
class(Vectors_ls)
```
\normalsize
\vspace{.4cm}
Keep in mind that a list **can contain every conceivable object** of `R` **within** every list position (yes, even lists of lists are possible).  
You can also **convert** object types to be of `list` by using the function `as.list()`.

### Matrix
`Matrices` are created using the function `matrix(data, nrow, ncol, byrow, dimnames)` ("data" represents the data; "nrow" and "ncol" indicate the number of rows and columns respectively, "byrow" is a logical argument indicating whether to fill the data into the matrix by row or by column, "dimnames" let's you ascribe names to columns and rows) in `R`. 
\vspace{.2cm}
\scriptsize
```{r createMatrices1}
Combine_mat <- matrix(data = c(Numbers_vec, Letters_vec), ncol = 2)
Combine_mat
class(Combine_mat)
```
\normalsize
\vspace{.4cm}
You can also **convert** object types to be of `matrix` by using the function `as.matrix()`.

### Data Frame I (creating from a `matrix`)
`Data Frames` are created using the function `data.frame(...)` ("..." can stand for a `matrix` or index individual columns) in `R`.  
\vspace{.4cm}

* Creating a `data.frame` from a `matrix`:  

\vspace{.2cm}
\scriptsize
```{r createDFs1}
Combine_df <- data.frame(Combine_mat)
Combine_df
class(Combine_df)
```
\normalsize
\vspace{.4cm}
You can also edit the names of columns and rows in a `data.frame` object using the commands `colnames()` and `rownames()` respectively.

### Data Frame II (creating with individual columns)
* Creating a `data.frame` from `vectors` as individual columns:  

\vspace{.2cm}
\scriptsize
```{r createDFs2}
Combine2_df <- data.frame(Numbers = Numbers_vec, Letters = Letters_vec)
Combine2_df
class(Combine_df)
```
\normalsize
\vspace{.4cm}
You can also **convert** object types to be of `data.frame` by using the function `as.data.frame()`.

## CheatSheet Overview

### Converting Modes
\scriptsize

| Target Mode | Function | What Happens |
|------:|:-----:|:-----|
| `numeric` | `as.numeric` | `FALSE` $->$ `0` |
|  |  | `TRUE` $->$ `1` |
|  |  | `"1", "2", "..."` $->$ `1, 2, ...` |
|  |  | `"A"` $->$ `NA`|
| `character` | `as.character` | `1, 2, ...` $->$ `"1", "2", "..."` |
|  |  | `FALSE` $->$ `"FALSE"` |
|  |  | `TRUE` $->$ `"TRUE"` |
| `logical` | `as.logical` | `0` $->$ `FALSE`  |
|  |  | any number that isn't `0` $->$ `TRUE` |
|  |  | `"FALSE"`/`"F" ` $->$ `FALSE` |
|  |  | `"TRUE"`/`"T" ` $->$ `TRUE` |
|  |  |  any character that isn't any of the above $->$ `NA`|

### Converting Types
\scriptsize

| Target Type | Function | What Happens |
|------:|:-----:|:-----|
| `vector` | `as.vector` | factor levels are dropped |
|  |  | matrices are made into vectors one column at a time  |
|  |  | data frames are made into vectors but remain in an array arrangement  |
|  |  | lists remain as lists  |
| `factor` | `as.factor` | factor levels are established from vector values |
|  |  | factors levels drawn from a column-wise vectorisation of matrices |
|  |  | does not work on data frames or lists |
| `matrix` | `as.matrix` | Puts content of vectors or factor into a single column |
|  |  | data frames remain unaltered as far as data structure goes |
|  |  | lists turn into cells with type and number of items of list positions |
| `data frame` | `as.data.frame` | each object gets put into a separate column |
|  |  | matrices remain unaltered as far as data structure goes |
|  |  | every list position is made into a column |
| `list` | `as.list` | every element gets put into an individual list position
|  |  | columns of data frames occupy list positions |

## Inspecting and Subsetting Objects in `R`
### Object Dimensions (`matrices` and `data.frames`)
Object **dimensions** are the last of the essential attributes of objects we will consider. They tell you how data is arranged. They are assessed using the `dim(...)` function in `R` ("..." stands for any given object name in `R`):  
\vspace{.4cm}

* Matrices have dimensions of rowcount `x` columncount

\vspace{.2cm}
\footnotesize
```{r, Dimensions3}
dim(Combine_mat)
```
\normalsize
\vspace{.4cm}

* Data Frames also have dimensions of rowcount `x` columncount

\vspace{.2cm}
\footnotesize
```{r, Dimensions4}
dim(Combine_df)
```
\normalsize

### Object Dimensions (`vectors`, `factors`, `lists`)


* Vectors, Factors and Lists don't have `dimensions`...  

\vspace{.2cm}
\footnotesize
```{r, Dimensions}
c(dim(Letters_vec), dim(Letters_fac), dim(Vectors_ls))
```
\normalsize
\vspace{.2cm}

* ... they only have a `length`

\footnotesize
\vspace{.2cm}
```{r, Dimensions2}
c(length(Letters_vec), length(Letters_fac), length(Vectors_ls))
```
\normalsize

### The Naming System (`data.frames`)
**Names** can serve as *labels* to elements of an object and are implemented/ assigned using the `names()` function at the most basic level (i.e. "for ``vectors` and `factors`"). Further implementations of *names* come in the form of the following functions:  
- `colnames()` (column labels, `data.frames`)  
- `rownames()` (row labels, `data.frames`)  
- `dimnames()` (dimension labels, `matrices`)  

\vspace{.2cm}
\scriptsize
```{r SubsetNames3}
colnames(Combine2_df)
rownames(Combine2_df)
Combine2_df$Numbers # subsetting by column 'Numbers'
```

### The Naming System (`vectors`)
* Our vectors don't have names assigned yet: 

\vspace{.2cm}
\scriptsize
```{r SubsetNames1}
names(Numbers_vec)
```
\normalsize
\vspace{.4cm}

* Let's do that:  

\vspace{.2cm}
\scriptsize
```{r SubsetNames2}
SubNames1 <- Numbers_vec
names(SubNames1) <- Letters_vec
SubNames1
```

### The Indexing System (`matrices` and `data.frames`)
The **indexing** system is basically the *numerical* **counterpart** to the naming system and called into action by the **use of square brackets** (`[]`):  
- `[elementnumber]` for `vectors and factors`  
- `[[elementnumber]]` for `lists`  
- `[rownumber, columnnumber]` for `data.frames` and `matrices`  

\vspace{.4cm}

* First element of second column of our matrix:

\vspace{.2cm}
\scriptsize
```{r Indexing1}
Combine_mat[1,2]
```
\normalsize 
\vspace{.4cm}

* Entire first column of our data frame:  

\vspace{.2cm}
\scriptsize
```{r Indexing2}
Combine_df[,1]
```

### The Indexing System (`vectors`, `factors` and `lists`)
* Third element of our letter vector:

\vspace{.2cm}
\scriptsize
```{r Indexing3}
Letters_vec[3]
```
\normalsize 
\vspace{.4cm}

* First element of our list:  

\vspace{.2cm}
\scriptsize
```{r Indexing4}
Vectors_ls[[1]]
```


# Functionality in `R`
## Vectorisation
### What is Vectorisation and why should I care?
\centering
`R` is a **vectorised** language.
\flushleft

\pause

\vspace{.4cm}

* **What** does that mean?  
Any mathematical operation is applied to every element in an object:  

\vspace{.2cm}
\small
```{r Vectorise1}
Numbers_vec + 1
```
\normalsize

\vspace{.4cm}

\pause

* **Why** care?  
Because it greatly influences how you do calculations in `R`.

## User-defined Functions
### Writing Functions in `R`
**Functions** are *special objects* within `R` as they **carry information for processing objects**. Functions require the following **components**:  
- *Name* - each function has a name  
- *Argument(s)* - give additional information to the function  
- *Call* - a function needs to be called to exert its effect  

\vspace{0.4cm}

Users can **create** their own **functions** in `R` as follows:

\vspace{.2cm}
\scriptsize
```{r Userfunction}
Plus1 <- function(x){ # function has an argument of `x`
  y <- x + 1 # 1 is added to the object `x` and saved as `y`
  return(y) # object `y` is returned
}
Plus1(Numbers_vec) # call the function on thes Numbers vector
```

\vspace{.2cm}

`#` indicate **comments in the code** - these are not run but can help explain stuff to the user

## Packages
### What are `R` packages?
**Packages** are `R`'s way of supplying the user with a widened **range of functionality** (just like a mod to a computer game or bonus tracks on a CD).  

\vspace{0.5cm}
\pause
There are **thousands of packges** for `R` which have been designed by other `R` users, tested vigorously, and are available freely for you to use.

\vspace{0.5cm}
\pause
All packages are available via the Comprehensive R Archive Network (\url{https://cran.r-project.org/}) and an overview of available packages can be retrieved here: \url{https://cran.r-project.org/web/packages/available_packages_by_date.html}.

### What do I do with packages?
\centering
You **install** and **load** them into `R`!  

\vspace{.4cm}
\flushleft
\pause

This is done in a **two-step process**:

\pause
\vspace{.4cm}

* `install.packages()` is used to install packages in `R`  

\vspace{.2cm}
\small
```{r, InstallP, eval=FALSE}
# vegan is a common library of functionality in biostatistics
install.packages("vegan") 
```
\normalsize
\pause
\vspace{.4cm}

* `library()` is used to load them into the current working environment  

\vspace{.2cm}
\small
```{r, LoadP}
library(vegan)
```

## Statements
### Logical statements
**Logical statements** are indicators of **whether something is true or not**. We use those frequently in real life (i.e. 'Is a 10 ECTS course worth more to me than a 5 ECTS course?'). `R` implements these with the following operators:  

| Operator | Translation |
|------:|:-----|
| `==` | "equals"  |
| `!=` | "does not equal"  |
| `<` | "is smaller"  |
| `<=` | "is equal or smaller"  |
| `>` | "is bigger"  |
| `>=` | "is equal or bigger" |

These statements return an element of mode `logical` (`TRUE` or `FALSE`).

### `if()` statements
`if()` statements *build on logical statements*. They **test whether something is correct** and then act on it:

\vspace{.2cm}
\small
```{r ifState}
# is 1 smaller than 2?
if(Numbers_vec[1] < Numbers_vec[2]){ 
  # if the statement is correct
  print("Is smaller") # print this to the console
}else{ 
  # if the statement is not correct
  print("Is not smaller") # print this to the console
}
```

## Loops
### `for()` loops
`for()` loops are in **action whilst** an **indiciator** is **within a specified data range**:

\vspace{.2cm}
\small
```{r forloop}
# loop from one to length of the letter vector in steps of 1
for(i in 1:length(Letters_vec)){
  # print current itteration element of letters vector
  print(Letters_vec[i])
}
```

### `while()` loops
`while()` loops work a lot like `for()` loops and are in **action whilst** an **indiciator statement** is `TRUE`:

\vspace{.2cm}
\small
```{r whileloop}
# while our data frame has equal to or less than 3 columns
while(dim(Combine_df)[2] <= 3){
  # bind letters factor vector to data frame as column
  Combine_df <- cbind(Combine_df, Letters_fac)
}
Combine_df # inspect the result
```

## Useful Commands In `R`
### `rm(list=ls())` I
* `rm(list=ls())` clears the entire working environment

\vspace{.2cm}
\begin{center}
  \includegraphics[width=0.8\linewidth]{Figures/R_RM.png}
\end{center}

### `rm(list=ls())` II

\centering `rm(list=ls())` is **extremely useful** and one should use this before sourcing (running an entire script) any `R` document to avoid influence of remnants of previous `R` sessions on the current analysis.

\vspace{.2cm}
\begin{center}
  \includegraphics[width=0.4\linewidth]{Figures/Meme_RMList.jpg}
\end{center}
\vspace{.2cm}
\pause
$\rightarrow$ Simply code this to be \textbf{your first line}!

### `getwd()` and `setwd()`
* `getwd()` returns the current working directory

$\rightarrow$ can identify the project folder if coded into script fairly early (usually second line)  

$\rightarrow$ can be used to soft code the working directory

\hfill\linebreak

\pause

* `setwd()` is used to set specific working directories

$\rightarrow$ can be used to address specific folders on your hard drive as the current working directory  

$\rightarrow$ often used to hard code the working directory (please avoid this at all cost!)

### `?`
* `?` followed by a function name, this will open the help documentation for said function

\vspace{.2cm}
\begin{center}
  \includegraphics[width=0.8\linewidth]{Figures/R_Fun.png}
\end{center}

### `View()`
* `View()` is a very neat command that lets the user inspect any object of their choosing in a separate tab

\vspace{.2cm}
\begin{center}
  \includegraphics[width=0.7\linewidth]{Figures/R_View1.png}
\end{center}

# Exercise
### Creating and Inspecting Objects (`vectors` and `factors`)
Create the following `vectors` or `factors`:

\footnotesize

| Name  | Content|
|------:|:------|
| `Letters_vec` | A vector reading "A", "B", "C" |
| `Numbers_vec` | A vector reading 1, 2, 3 |  
| `Logic_vec` | A vector reading `TRUE`, `FALSE` |
| `Big_vec` | A vector of the elements of the first three vectors |
| `Seq_vec` | A vector reading as a sequence of full numbers from 1 to 20 |
| `Letters_fac` | A factor reading "A", "B", "C" |
| `Numbers_fac` | A factor reading 1, 2, 3 |
| `Constrained_fac` | A factor reading 1, 2, 3, levels 1 and 2 are allowed |
| `Expanded_fac` | A factor reading 1, 2, 3 levels 1 - 4 are allowed |

\normalsize
Inspect these objects for their classes and dimensions!

### Creating and Inspecting Objects (`matrices`, `data.frames` and `lists`)
Create the following objects:

\footnotesize

| Name  | Content|
|------:|:------|
| `Combine_mat` | `Numbers_vec` and `Letters_vec` in columns of a matrix |
| `Pivot_mat` | First two vectors in distinct rows of a matrix |
| `Names_mat` | The above matrix with meaningful names |
| `Combine_df` | The first matrix we established as a data frame |
| `Names_df` | The previous data frame with meaningful names |
| `Vectors_ls` | The first two vectors we created as a list |

\normalsize
Inspect these objects for their classes and dimensions!

### Statements and Loops
Test the following *statements*:  

* `Numbers_vec` contains more elements than `Letters_fac`  
* The first column of `Combine_df` is shorter than `Vectors_ls`  
* The elements of `Letters_vec` are the same as the elements of `Letters_fac`  

\vspace{0.2cm}
Write the following *loops*:  

* Print each element of `Vectors_ls`  
* Print each element of `Numbers_vec` + 1  
* Subtract 1 from each element of the first column of `Combine_mat` and print each element separately

### Using the Useful Commands
Do the following:  

* Read out your current working directory  
* Inspect the `Vectors_ls` object using the `View()` function  
* Inspect the `Combine_df` object using the `View()` function  
* Get the help documentation for the `as.matrix()` function  
* Install and load the `dplyr` package  
* Remove the `Logic_vec` object from your working environment  
* Clear your entire working environment  