---
title: "Nominal Tests"
# subtitle: "And Introduction to the Course and The Importance of Proper Statistics"
author: "Erik Kusch"
date: ""
fontsize: 10pt
output: 
  beamer_presentation: # handouts can be established by setting classoption: "handout" instead of classoption: t
    keep_tex: true
    toc: false # this is added through a later command
    slide_level: 3 # at how many pound signs (#) to assume slide title level
    includes:
      in_header: Style.tex
classoption: "handout"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=TRUE, tidy.opts=list(width.cutoff=50),tidy=TRUE, fig.height=8)
options(digits=2, width = 60)

# function to load packages and install them if they haven't been installed yet
install.load.package <- function(x) {
  if (!require(x, character.only = TRUE))
    install.packages(x)
  require(x, character.only = TRUE)
}
package_vec <- c("nonpar" # needed for Cochran's Q
                 )
sapply(package_vec, install.load.package)
```

<!--- ####### TOC (use this if you want ## headers to be included in the toc)--------------------
---------------------------------------  --->

\tableofcontents[subsubsectionstyle=hide]

<!--- ####### Inferential Statistics-------------------------------------------------------------
---------------------------------------  --->
# Background
### Introduction
\begin{center}
These approaches only allow for the use of \textbf{categorical} (nominal) \textbf{variables}!
\end{center}
\vspace{.5cm}
\pause
Prominent nominal tests include:
\begin{itemize}
\item \textbf{Binomial Test}
\item \textbf{McNemar}
\item Fisher's Exact
\item \textbf{Cochran's Q}
\item \textbf{Chi-Squared}
\item ...
\end{itemize}


### The `table()` function
In reality, you will need to convert your data to fit the various nominal test specifications. To do so, you may wish to enlist the help of the `table()` function of base `R` which converts nominal records into count data.
\vspace{.5cm}
\scriptsize
```{r Table}
Samples <- c("A", "B")
set.seed(42)
counts <- sample(Samples, size = 1000, replace = TRUE)
table(counts)
```

# Analyses
## Binomial Test
### Purpose And Assumptions
\textbf{Binomial Test} \flushright `binom.test()` in base `R`  
\vspace{-10pt}
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} &  To test whether the observed distribution of data values of a binomial variable differ from what was expected. \vspace{5pt} \\
    \pause
    $H_0$ & \textit{The observed binomial data proportions do not differ significantly from the expected proportions.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Variable values are binomial.
      \item The population is significantly larger than the sample.
      \item The sample accurately represents the population.
      \item Sampled values are independent (one value does not influence another).
      \end{itemize}
     \\
  \end{tabular}}
\end{table}

### Minimal Working Example
We feed the `binom.test()` function a 1/4 (`c(200, 800)`) data set whilst expecting the distribution to be 1 to 1 (`p = 0.5`).  

\scriptsize
\vspace{.2cm}
```{r Binomial}
binom.test(c(200, 800), p = 0.5)
```
\normalsize
\vspace{.2cm}
The result is significant (p $\approx$ `r round(binom.test(c(200, 800), p = 0.5)$p.value,2)`).

## McNemar
### Purpose And Assumptions
\textbf{McNemar} \flushright `mcnemar.test()` in base `R`  
\vspace{-10pt}
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} &  To test whether there is a change in proportion of paired data.  \vspace{5pt} \\
   \pause
    $H_0$ & \textit{The observed binomial data proportions do not differ significantly between treatments/paired sets.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Variable values are binomial.
      \item The population is significantly larger than the sample.
      \item The sample accurately represents the population.
      \end{itemize}
     \\
  \end{tabular}}
\end{table}

### Minimal Working Example
We feed the `mcnemar.test()` function a 1 to 1 (`c(500, 500)`) as well as a a 1 to 4 (`c(200, 800)`) data set for the paired data sets.  
\vspace{.2cm}
\scriptsize
```{r McNemar}
Performance <- matrix(c(500, 500, 200, 800),nrow = 2)
Performance
mcnemar.test(Performance)
```
\normalsize
\vspace{.2cm}
With a p-value of $\approx$ `r round(mcnemar.test(Performance)$p.value,2)` the test concludes significantly.


## Cochran's Q
### Purpose And Assumptions
\textbf{Cochran's Q} \flushright `cochrans.q()` in the `nonpar` package  
\vspace{-10pt}
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} &  To test whether there are differences in matched sets of three or more frequencies or proportions. \vspace{5pt}\\
    \pause
    $H_0$ & \textit{The observed proportions of data values in treatments is equal among the matched sets.} \\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item The first (dependent/response) variable is binomial.
      \item The second (independent/predictor) variable is nominal/categorical with three values.
      \item The population is significantly larger than the sample.
      \item The sample accurately represents the population.
      \end{itemize}
     \\
  \end{tabular}}
\end{table}

### Minimal Working Example
We feed the `cochrans.q()` function a 6 by 4 matrix of binomial values. The first column represents our dependent, binomial variable. The remaining columns represent our independent variable on three levels expressed as binomial values.  
\vspace{.2cm}
\tiny
```{r CochransQ}
CochranMatrix <- matrix(c(1,1,1,1,1,1,
              1,1,0,1,1,1,
              0,0,0,1,0,0,
              0,1,0,0,1,1), 6, 4)
cochrans.q(CochranMatrix)
```

## Chi-Squared
### Purpose And Assumptions
\textbf{Chi-Squared} \flushright `chisq.test()` in base `R`  
\vspace{-10pt}
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} &  To tests whether distributions of categorical variables differ from one another thus identifying whether they are related. \vspace{5pt}\\
    \pause
    $H_0$ & \textit{The distributions of nominal variables are equal.} \\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Variable values are nominal/categorical.
      \item The population is significantly larger than the sample.
      \item The sample accurately represents the population.
      \item Sampled values are independent (one value does not influence another).
      \end{itemize}
     \\
  \end{tabular}}
\end{table}

### Minimal Working Example - One Sample Situation I
We feed the `chisq.test()` function an unbiased nominal set of three levels:
\vspace{.2cm}
\tiny
```{r ChiSquared1}
set.seed(42)
ChiMat1 <- table(sample(c("A", "B", "C"), 1000, replace = TRUE))
ChiMat1
chisq.test(ChiMat1)
```
\vspace{.2cm}
\normalsize
Obviously, the observed distribution does not differ from our expectation of equally distributed proportions and so the test concludes non-significantly.

### Minimal Working Example - One Sample Situation II
We feed the `chisq.test()` function a skewed (towards `"A"`) nominal set of three levels:

\tiny
\vspace{.2cm}
```{r ChiSquared2}
set.seed(42)
ChiMat2 <- table(sample(c("A", "B", "C"), 1000, replace = TRUE, prob = c(0.8, 0.1, 0.1)))
ChiMat2
chisq.test(ChiMat2)
```
\vspace{.2cm}
\normalsize
Obviously, the observed distribution does differ from our expectation of equally distributed proportions and so the test concludes significantly.

### Minimal Working Example - Two Sample Situation
We feed the `chisq.test()` our unbiased as well as our skewed (towards `"A"`) nominal set of three levels to see whether their distributions differ significantly.
\vspace{.2cm}
\tiny
```{r ChiSquared3}
ChiMatrix <- cbind(ChiMat1, ChiMat2)
ChiMatrix
chisq.test(ChiMatrix)
```
\vspace{.2cm}
\normalsize
Clearly, they do differ significantly.


# Our Data
## Choice Of Variables
### Variables We Can Use
\begin{center}
Which variables in our \textit{Passer domesticus} data set are nominal?
\end{center}

\vspace{.8cm}

\begincols[T]
\begincol{.5\linewidth}
\begin{itemize}
\visible<2->{
\item Site Index
}
\visible<3->{
\item Climate
}
\visible<4->{
\item Population Status
}
\visible<5->{
\item Colour
}
\visible<6->{
\item Sex
}
\end{itemize}
\endcol
\begincol{.5\linewidth}
\begin{itemize}
\visible<7->{
\item Nesting Site
}
\visible<8->{
\item Flock
}
\visible<9->{
\item Home Range
}
\visible<10->{
\item Predator Presence
}
\visible<11->{
\item Predator Type
}
\end{itemize}
\endcol
\endcols

\vspace{20pt}
\vspace{.5cm}
\visible<12>{
\begin{center}
All of these are nominal but some are binomial.
\end{center}
}

## Research Questions
### Research Questions And Hypotheses
\begin{center}
So which of our major research questions (seminar 6) can we answer?
\end{center}
\vspace{.4cm}
\small
\begincols[T]
  \begincol{.5\linewidth}
  \pause
    \textbf{Binomial Test}
    \begin{itemize}
    \item \textit{Sexual Dimorphism:} Are the sexes represented in equal proportions? 
    \item \textit{Predation:} Are our sites dominated by predators or not?
    \end{itemize} 
  \pause
    \textbf{McNemar}
    \begin{itemize}
    \item \textit{Sexual Dimorphism:} Compare sex ratio over time (we need to generate some new data for this using the sample function).
    \end{itemize} 
\endcol
\begincol{.5\linewidth}
  \pause
    \textbf{Cochran's Q}
    \begin{itemize}
    \item \textit{Sexual Dimorphism:} Are colours related to sex?
    \item \textit{Predation:} Are colours related to predators?
    \end{itemize} 
  \pause
    \textbf{Chi-Squared}
    \begin{itemize}
    \item \textit{Sexual Dimorphism:} Are colours related to sex?
    \item \textit{Predation:} Are colours or nesting sites related to predators?
    \end{itemize} 
\endcol
\endcols
\vspace{.4cm}
\raggedright Use the `1 - Sparrow_Data_READY.rds` data set for these analyses.