---
title: "Metric Tests (More-Than-Two-Sample Situations)"
# subtitle: "And Introduction to the Course and The Importance of Proper Statistics"
author: "Erik Kusch"
date: "??/??/20??"
fontsize: 10pt
output: 
  beamer_presentation: # handouts can be established by setting classoption: "handout" instead of classoption: t
    keep_tex: false
    toc: false # this is added through a later command
    slide_level: 3 # at how many pound signs (#) to assume slide title level
    includes:
      in_header: Style.tex
classoption: t
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=TRUE, tidy.opts=list(width.cutoff=50),tidy=TRUE, fig.height=8)
options(digits=2, width = 60)

# function to load packages and install them if they haven't been installed yet
install.load.package <- function(x) {
  if (!require(x, character.only = TRUE))
    install.packages(x)
  require(x, character.only = TRUE)
}
package_vec <- c("nonpar" # needed for Cochran's Q
                 )
sapply(package_vec, install.load.package)
```

<!--- ####### TOC (use this if you want ## headers to be included in the toc)--------------------
---------------------------------------  --->

\tableofcontents[subsubsectionstyle=hide]

<!--- ####### Inferential Statistics-------------------------------------------------------------
---------------------------------------  --->
# Background
### Introduction
\begin{center}
Metric tests are used to compare parameters of metric/ordinal variable values among groups/individuals.
\end{center}
\pause
Prominent metric tests for more-than-two-sample situations include:
\begin{itemize}
\item \textbf{Kruskal-Wallis Test}
\item \textbf{Friedman Test}
\item t Test (dealt with in Seminar 12)
\item ...
\end{itemize}

### Terminology

Remember:
\begin{itemize}
\item \textbf{Got two samples?}
  \begin{itemize}
  \item \textit{Independent.} $\rightarrow$ Mann-Whitney U Test
  \item \textit{Dependent.} $\rightarrow$ Wilcoxon Signed Rank Test
  \end{itemize}
\item \textbf{Got more than two samples?}
  \begin{itemize}
  \item \textit{Independent.} $\rightarrow$ Kruskal-Wallis Test
  \item \textit{Dependent.} $\rightarrow$ Friedman Test
  \end{itemize}
\end{itemize}

# Analyses
## Kruskal-Wallis Test
### Purpose And Assumptions
\textbf{Kruskal-Wallis Test} \flushright `kruskal.test()` in base `R`  
\vspace{-10pt}
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} &  To identify whether groups of variable values are different from one another. \vspace{5pt} \\
    \pause
    $H_0$ & \textit{There is no difference in characteristics of the response variable values in dependence of the classes of the predictor variable.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Predictor variable is categorical (not binary!)
      \item Response variable is ordinal or metric
      \item Variable values are \textbf{independent} (not paired)
      \end{itemize}
     \\
  \end{tabular}}
\end{table}

### Minimal Working Example
Let's use the `kruskal.test()` function to test whether the medians of an unnamed variable of three unconnected populations (`a`, `b` and `c`) with 10 individuals each are truly different:
\scriptsize
```{r Kruskal}
set.seed(42)
a <- rnorm(n = 10, mean = 15, sd = 3)
b <- rnorm(n = 10, mean = 10, sd = 3)
c <- rnorm(n = 10, mean = 5, sd = 3)
groups <- as.factor(rep(c("a", "b", "c"), each = 10))

kruskal.test(x = c(a, b, c), g = groups)
```

## Friedman Test
### Purpose And Assumptions
\textbf{Friedman Test} \flushright `friedman.test()` in base `R`  
\vspace{-10pt}
\begin{table}[c]
  \centering
  \resizebox{\textwidth}{!}{%
  \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
    \textit{Purpose:} &  To identify whether groups of variable values are different from one another. \vspace{5pt} \\
    \pause
    $H_0$ & \textit{There is no difference in characteristics of the response variable values in dependence of the classes of the predictor variable.}\\
    \pause
    \textit{Assumptions:} & 
      \begin{itemize}
      \item Predictor variable is categorical (not binary!)
      \item Response variable is ordinal or metric
      \item Variable values are \textbf{dependent} (paired)
      \end{itemize}
     \\
  \end{tabular}}
\end{table}

### Minimal Working Example
Let's use the `friedman.test()` function to test whether the medians of an unnamed variable of three connected samples (`a`, `b` and `c`) with 10 individuals each (i.e. one re-sampled population) are truly different:
\scriptsize
```{r Friedman}
set.seed(42)
a <- rnorm(n = 10, mean = 15, sd = 3)
b <- rnorm(n = 10, mean = 10, sd = 3)
c <- rnorm(n = 10, mean = 5, sd = 3)
TestData <- matrix(c(a,b,c), nrow = 10, byrow = FALSE, dimnames = list(1:10, c("a", "b", "c")))

friedman.test(y = TestData)
```

# Our Data
## Choice Of Variables
### Variables We Can Use

\begincols[T]
  \begincol{.5\linewidth}
    \textbf{Response variables} (metric/ordinal)
    \visible<2->{
    \begin{itemize}
    \item Weight
    \item Height
    \item Wing Chord
    \item Nesting Height
    \item Number of Eggs
    \item Egg Weight
    \item Home Range
    \end{itemize} 
    }
\endcol
\begincol{.5\linewidth}
  \pause \pause
    \textbf{Predictor variables} (categorical but not binary)
    \visible<4->{
    \begin{itemize}
    \item Home Range (3 levels - Small, Medium, Large)
    \item Site Index (11 levels)
    \item Predator Presence/Type (3 levels - Avian vs. Non-Avian vs. None)
    \item Climate (3 levels - Continental, Semi-Coastal, Coastal)
    \end{itemize} 
    }
\endcol
\endcols

## Research Questions
### Research Questions And Hypotheses
\begin{center}
So which of our major research questions (seminar 6) can we answer?
\end{center}

\small
\begincols[T]
  \begincol{.5\linewidth}
  \pause
    \textbf{Kruskal-Wallis Test}
    \begin{itemize}
    \item \textit{Site-wise variation:} Do characteristics of \textit{Passer domesticus} depend on the site they are recorded at?
    \item \textit{Climate Warming/Extremes:} Does sparrow morphology depend on climate?
    \item \textit{Predation:} Does nesting height depend on predator characteristics?
    \item \textit{Competition:} Does home range depend on climate?
    \end{itemize} 
\endcol
\begincol{.5\linewidth}
  \pause
    \textbf{Friedman Test} (suppose a resettling program)
    \begin{itemize}
    \item \textit{Climate Warming/Extremes:} Does sparrow morphology change depend on climate?
    \item \textit{Predation:} Does nesting height depend on predator characteristics?
    \item \textit{Competition:} Does home range depend on climate?
    \end{itemize} 
    \footnotesize
    $\rightarrow$ Find the data in \linebreak \textit{2a - Sparrow\_ResettledSIMA\_READY.rds} \linebreak and 
    \textit{2b - Sparrow\_ResettledSIUK\_READY.rds}
\endcol
\endcols

# Exercise

### Writng A Function In `R`
\begin{center}
Establishing \textit{user-defined} functions is at the \textbf{heart of `R`}!
\end{center}

A function requires:  
+ A **name**  
+ **Arguments**  
+ to be **called**
```{r Func}
Fun <- function(argument){
  print(argument)
}
Fun(argument = "Test")
```