---
title: "A Primer For Statistical Tests"
# subtitle: "And Introduction to the Course and The Importance of Proper Statistics"
author: "Erik Kusch"
date: ""
fontsize: 10pt
output: 
  beamer_presentation: # handouts can be established by setting classoption: "handout" instead of classoption: t
    keep_tex: true
    toc: false # this is added through a later command
    slide_level: 3 # at how many pound signs (#) to assume slide title level
    includes:
      in_header: Style.tex
classoption: "handout"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=TRUE, tidy.opts=list(width.cutoff=20),tidy=TRUE, fig.height=8)
options(digits=2)
```

<!--- ####### TOC (use this if you want ## headers to be included in the toc)--------------------
---------------------------------------  --->

\tableofcontents[subsubsectionstyle=hide]

<!--- ####### Descriptive Statistics-------------------------------------------------------------
---------------------------------------  --->
# Variables
## What Are Variables?
### Variables And Their Subsets
\textbf{What is a variable?}    
A \textit{variable} presents more or less valuable information about a potential multitude of characteristics of a study system.

\vspace{0.4cm}
\pause

\textbf{What's the fuss?}  
Sampling effort is limited. We only ever sample a subset of globally present values of any given variable. 

\vspace{0.4cm}
\pause

\textbf{So?}  
Every variable is subject to a certain distribution of its values and each measurement is expected to follow the global distribution to a certain degree.

\vspace{0.4cm}
\pause
\begin{tcolorbox}[colback=bondiblue!5,colframe=bondiblue!40!black,title=]
\centering Variables are, more or less, \textbf{raw data}.
\end{tcolorbox}

## Types of Variables

### Types of Variables
Variables can be classed into a multitude of types. The most common classification system knows:
\vspace{0.6cm}
\begin{columns}[T]
  \begin{column}{.5\linewidth}
  \pause
    \textbf{Categorical Variables}
    \begin{itemize}
    \item also known as \textit{Qualitative Variables}
    \item Scales can be either:
      \begin{itemize}
      \item Nominal
      \item Ordinal
      \end{itemize}
    \end{itemize}
  \end{column}
  \begin{column}{.5\linewidth}
  \pause
    \textbf{Continuous Variables}
    \begin{itemize}
    \item also known as \textit{Quantitative Variables}
    \item Scales can be either:
      \begin{itemize}
      \item Discrete
      \item Continuous
      \end{itemize}
    \end{itemize}
  \end{column}
\end{columns}



### Categorical Variables
\begin{tcolorbox}[colback=bondiblue!5,colframe=bondiblue!40!black,title=]
\centering Categorical variables are those variables which \textbf{establish and fall into distinct groups and classes}.
\end{tcolorbox}
\pause
\vspace{0.4cm}
Categorical variables:
\begin{itemize}
\item can take on a finite number of values
\item assign each unit of the population to one of a finite number of groups
\item can \textit{sometimes} be ordered
\end{itemize}
\pause
\vspace{0.4cm}
\textbf{In R}, categorical variables usually come up as object type `factor` or `character`.

### Categorical Variables (Examples)
Examples of categorical variables:
\vspace{.5cm}
\begin{itemize}
\pause
\item Biome Classifications (e.g. "Boreal Forest", "Tundra", etc.)
\pause
\item Sex (e.g. "Male", "Female")
\pause
\item Hierarchy Position (e.g. "$\alpha$-Individual", "$\beta$-Individual", etc.)
\pause
\item Soil Type (e.g. "Sandy", "Mud", "Permafrost", etc.)
\pause
\item Leaf Type (e.g. "Compound", "Single Blade", etc.)
\pause
\item Sexual Reproductive Stage (e.g. "Juvenile", "Mature", etc.)
\pause
\item Species Membership
\pause
\item Family Group Membership
\pause
\item ...
\end{itemize}

### Continuous Variables
\begin{tcolorbox}[colback=bondiblue!5,colframe=bondiblue!40!black,title=]
\centering Continuous variables are those variables which \textbf{establish a range of possible data values}.
\end{tcolorbox}
\pause
\vspace{0.4cm}
Continuous variables:
\begin{itemize}
\item can take on an infinite number of values
\item can take on a new value for each unit in the set-up
\item can \textit{always} be ordered
\end{itemize}
\pause
\vspace{0.4cm}
\textbf{In R}, continuous variables usually come up as object type `numeric`.

### Continuous Variables (Examples)
Examples of categorical variables:
\vspace{.5cm}
\begin{itemize}
\pause
\item Temperature
\pause
\item Precipitation
\pause
\item Weight
\pause
\item pH
\pause
\item Altitude
\pause
\item Group Size
\pause
\item Vegetation Indices
\pause
\item Time
\pause
\item ...
\end{itemize}

### Converting Variable Types
\textit{Continuous variables} can be converted into \textit{categorical variables} via a method called \textbf{binning:}

\vspace{0.1cm}

\pause
Given a variable range, one can establish however many "bins" as one wants. For example:
\pause
\begin{itemize}
\item Given a temperature range of $271K - 291K$, there may be 4 bins of equal size:
  \begin{itemize}
  \item Bin A: $271K \leq X \leq 276K$
  \item Bin B: $276K < X \leq 281K$
  \item Bin C: $281K < X \leq 286K$
  \item Bin D: $286K < X \leq 291K$
  \end{itemize}

\end{itemize}
\vspace{0.2cm}
\pause
\begin{tcolorbox}[colback=burgundy!5,colframe=burgundy!40!black,title=]
\centering Whilst a \textbf{continuous variable} can be both \textit{continuous} and \textit{categorical}, a \textbf{categorical variable} can only ever be \textit{categorical}!
\end{tcolorbox}

## Variables And Scales
### Variables On Scales
Another way of classifying variables are the \textbf{scales} they are represented on.
\vspace{0.4cm}
\pause
\begin{tcolorbox}[colback=burgundy!5,colframe=burgundy!40!black,title=]
\centering \textbf{Different scales} of variables \textbf{require different statistical procedures} for analyses!
\end{tcolorbox}
\pause 
\vspace{0.4cm}
Variable scales include:
\begincols[T]
  \begincol{0.5\linewidth}
  \begin{itemize}
  \item \textbf{Nominal} 
  \item \textbf{Binary}
  \item \textbf{Ordinal}
  \end{itemize}
  \endcol
  \begincol{0.5\linewidth}
  \begin{itemize}
  \item \textbf{Interval}
  \item \textbf{Relation/Ratio}
  \end{itemize}
  \endcol
\endcols
\pause 
\vspace{0.4cm}
Some statistics books teach \textit{integer scales} along the above mentioned scales. Some people dispute this and claim these scales to be \textit{ratio scales}.

### Nominal And Binary
\textbf{Nominal scales} of variables correspond to \textit{categorical variables} which cannot be put into a meaningful order.
\begin{columns}[T]
  \begin{column}{.7\linewidth}
    \vspace{0.2cm}
    \pause
    \begin{itemize}
    \item Variables on nominal scales put units into distinct categories
    \item These variables may be numerical but offer no mathematical interpretation
    \end{itemize}
    \pause
    \textit{Examples:}
    \begin{itemize}
    \item Petal colour (red, green, blue, etc.)
    \item Individual IDs
    \end{itemize}
  \end{column}
  \begin{column}{.3\linewidth}
    \begin{center}
    \includegraphics[width=0.65\linewidth]{Figures/Nominal.png}  
    \end{center}
  \end{column}
\end{columns}
\vspace{0.2cm}
\pause
\textbf{Binary scales} are a special case of \textit{nominal scales} taking only two possible values: 0 and 1.

### Ordinal
\textbf{Ordinal scales} of variables correspond to \textit{categorical variables} which can be put into meaningful order.
\begin{columns}[T]
  \begin{column}{.7\linewidth}
    \vspace{0.2cm}
    \pause
    \begin{itemize}
    \item Variables on ordinal scales put units into distinct categories
    \item These variables may be numerical and mathematical interpretation
    \end{itemize}
    \pause
    \textit{Examples:}
    \begin{itemize}
    \item Size (small, medium, large, etc.)
    \item Binned continuous variables
    \end{itemize}
  \end{column}
  \begin{column}{.3\linewidth}
    \vspace{0.2cm}
    \begin{center}
    \includegraphics[width=0.65\linewidth]{Figures/Ordinal.png}  
    \end{center}
  \end{column}
\end{columns}

### Interval/Discrete
\textbf{Interval scales} of variables correspond to a mix of \textit{continuous variables}.
\begin{columns}[T]
  \begin{column}{.7\linewidth}
    \vspace{0.2cm}
    \pause
    \begin{itemize}
    \item Variables on interval scales are measured on equal intervals from a defined zero point/point of origin
    \item The point of origin \textbf{does not imply an absence of the measured characteristic}
    \end{itemize}
    \pause
    \textit{Examples:}
    \begin{itemize}
    \item Temperature [$\degree C$]
    \item pH
    \end{itemize}
  \end{column}
  \begin{column}{.3\linewidth}
    \vspace{0.2cm}
    \begin{center}
    \includegraphics[width=0.65\linewidth]{Figures/Interval.png}  
    \end{center}
  \end{column}
\end{columns}

### Relation/Ratio
\textbf{Relation/Ratio scales} of variables correspond to \textit{continuous variables}.
\begin{columns}[T]
  \begin{column}{.7\linewidth}
     \vspace{0.2cm}
    \pause
    \begin{itemize}
    \item Variables on relation/ratio scales are measured on equal intervals from a defined zero point/point of origin
    \item The point of origin \textbf{does imply an absence of the measured characteristic}
    \end{itemize}
    \pause
    \textit{Examples:}
    \begin{itemize}
    \item Temperature [$K$]
    \item Weight
    \end{itemize}
  \end{column}
  \begin{column}{.3\linewidth}
    \begin{center}
    \includegraphics[width=0.65\linewidth]{Figures/Relation.png}  
    \end{center}
  \end{column}
\end{columns}
\pause
\vspace{0.5cm}
\textbf{Integer scales} are a special case of \textit{ratio scales} allowing only for integral numbers.

### Confusion Of Units
\begin{center}
\includegraphics[width=0.5\linewidth]{Figures/Metrics.jpg}  
\end{center}

### Checklist For Your Variables
\begin{itemize}
\item \textbf{What variables} am I using and:
  \begin{itemize}
  \pause
  \item Of what \textit{mode} are they?
  \pause
  \item Is every record of the same variable based on the same \textit{unit of measure}?
  \pause
  \item Should I \textit{convert} some of them? \linebreak
  \end{itemize}
\pause
\item \textbf{What scales} apply and:
  \begin{itemize}
  \pause 
  \item What do they \textit{imply}?
  \pause 
  \item Does my data \textit{fit} these scales? \linebreak
  \end{itemize}
\end{itemize}
\pause
$\rightarrow$ You should be able to \textbf{answer these question \textit{before} you begin data collection!}




# Distributions
## The Basics of Distributions
### What Are Distributions?
\begincols[c]
  \begincol{0.5\linewidth}
  \begin{tcolorbox}[colback=bondiblue!5,colframe=bondiblue!40!black,title=]
  \centering A distribution of a statistical data set (sample/population) shows all the possible values/intervals of the data in question and their frequency.
  \end{tcolorbox}
  \endcol
  \begincol{0.5\linewidth}
  \visible<2->{
  \begin{center}
  \includegraphics[width=0.9\linewidth]{Figures/Distris.png} 
  \end{center}
  }
  \endcol
\endcols
\pause
\vspace{1cm}
$\rightarrow$ Basically \textbf{data patterns} we are considering/looking for.


### Frequency Distributions
\begincols[T]
  \begincol{0.5\linewidth}
  \textbf{Frequency} Distributions:
  \begin{itemize}
  \item \textit{Theory}
    \begin{itemize}
    \item Simple representations of data value frequencies
    \item Can be established for every variable
    \end{itemize}
    \pause
  \item \textit{Practice in R}
    \begin{itemize}
    \item Visualisation via the `hist()` function
    \end{itemize}
  \end{itemize}
\endcol
\begincol{0.5\linewidth}
```{r freqdist_plot, echo=FALSE}
hist(rnorm(100000,20,2),
main = "Frequency Distribution")
```
\endcol
\endcols
```{r freqdist, eval=FALSE, tidy.opts=list(width.cutoff=55), tidy=FALSE}
hist(rnorm(100000,20,2),
main = "Frequency Distribution")
```

### Probability Density Distributions I
\begincols[T]
  \begincol{0.5\linewidth}
  \textbf{Probability Density} Distributions:
  \begin{itemize}
  \item \textit{Theory}
    \begin{itemize}
    \item Representation of data value probabilities
    \item Can be established for \textit{continuous} variables
    \end{itemize}
    \pause
  \item \textit{Practice in R}
    \begin{itemize}
    \item Visualisation via the `density()` function
    \end{itemize}
  \end{itemize}
\endcol
\begincol{0.5\linewidth}
```{r densitydist_plot, echo=FALSE}
plot(density(rnorm(100000,20,2)),
main = "Probability Density Distribution")
```
\endcol
\endcols
```{r densitydist, eval=FALSE, tidy.opts=list(width.cutoff=55), tidy=FALSE}
plot(density(rnorm(100000,20,2)),
  main = "Probability Density Distribution")
```

### Probability Density Distributions II
\begin{tcolorbox}[colback=bondiblue!5,colframe=bondiblue!40!black,title=]
\centering Probability Density Distributions hold the \textbf{majority of importance} in statistics!
\end{tcolorbox}
\vspace{0.4cm}
\pause
A few key points about these distributions:
\begin{itemize}
\pause
\item Area under the curve (AUC) sums to 1
\pause
\item A probability for every given single value is 0
\pause
\item The AUC between two values on the X-axis equals the probability to randomly sample a value between these two points
\end{itemize}

## Normality
### Univariate Standard Normal/Gaussian Distribution
\begin{tcolorbox}[colback=bondiblue!5,colframe=bondiblue!40!black,title=]
\centering One of the \textbf{most important} distributions in natural sciences. 
\end{tcolorbox}

\begincols
  \begincol{.5\linewidth}
  \vspace{-0.2cm}
  \begin{itemize}
  \pause
  \item Used to represent real-valued random variables whose distributions are not known
  \pause
  \item The \textbf{central limit theorem} applies (draw a sufficient number of samples and you end up with the normal distribution)
  \pause
  \item These distributions are usually known also as "bell curves" (\textbf{Attention:} other distributions take this shape too)
  \end{itemize}
  \endcol
  \begincol{.5\linewidth}
  \visible<4->{
  \begin{center}
  \includegraphics[width=0.6\linewidth]{Figures/Normal.jpg}  
  \end{center}
  }
  \endcol
\endcols

### Testing For Normality
\begin{tcolorbox}[colback=burgundy!5,colframe=burgundy!40!black,title=]
\centering Testing for normality of the data is \textbf{crucial} for certain statistical procedures.
\end{tcolorbox}

\begincols[T]
\begincol{.5\linewidth}
  \visible<2->{\textbf{The Shapiro-Wilks Test In Theory}}
  \begin{itemize}
  \visible<3->{
  \item Base assumption: The data is normally distributed
  \item If $p$-value $<$ chosen significance level, the data is \textbf{not} normally distributed
  \item Very sensitive to sample size
  }
  \end{itemize}
\endcol
\begincol{.5\linewidth}
  \visible<2->{\textbf{The QQ Plot In Theory}}
  \begin{itemize}
  \visible<4->{
  \item Method for comparing two probability distributions by plotting their quantiles against each other
  \item If the two distributions being compared are similar, the plot will show the line $y = x$.
  \item Compare the data distribution to the normal distribution
  }
  \end{itemize}
\endcol
\endcols


### The Shapiro-Wilks Test In R
Using the `shapiro.test()` function:

\footnotesize
\begincols[T]
\pause
\begincol{.5\linewidth}
```{r shapiroyes}
shapiro.test(rnorm(5000,20,2))
```
$\rightarrow$ Clearly a normal distributed set of values
\endcol
\begincol{.5\linewidth}
\pause
```{r shapirono}
shapiro.test(seq(1,500,5))
```
$\rightarrow$ Clearly no normal distributed set of values
\endcol
\endcols


### The Q-Q Plot
Using the `qqnorm()` function:
\footnotesize
\begincols
\begincol{.5\linewidth}
\pause
```{r QQyes, tidy=FALSE}
qqnorm(rnorm(5000,20,2))
```
$\rightarrow$ Clearly a normal distributed set of values
\endcol
\begincol{.5\linewidth}
\pause
```{r QQno, tidy=FALSE}
qqnorm(seq(1,500,5))
```
$\rightarrow$ Clearly no normal distributed set of values
\endcol
\endcols


## What Distributions To Consider
### An Overview of Distributions
There is a \textbf{plethora of distributions} which variables could fall onto:
\vspace{.3cm}
    \begin{itemize}
    \pause
    \item Bernoulli (probabilities of value 1 and 0 are interdependent)
    \pause
    \item \textbf{Binomial} (number of successes in a series)
    \pause
    \item \textbf{Poisson} (probability of a given number of events occurring in a fixed interval of time or space)
    \pause
    \item Beta (family of two-parameter distributions with one mode)
    \pause
    \item Kent (three-dimensional sphere distribution)
    \pause
    \item \textbf{Univariate Standard Normal/Gaussian}
    \pause
    \item Multivariate Normal
    \pause
    \item Log-Normal
    \end{itemize}
\pause
\flushright ... and yet \textbf{the hat goes deeper}

### Binomial Distribution
One of the **more important** distributions. It is applicable to:  

\begincols[T]
  \begincol{.5\linewidth}
  \pause
  
  * Variables which can only take two possible values (e.g. "states")  
  * All records of the variable have the same probability *p* of being in one of the two states  
  
  \pause
  It is made up of three **criteria**:  
  \pause
  
  \begin{itemize}
  \item $p$ - the "success" probability  
  \item $n$ - sample size (how often we sample)  
  \item $N$ - the "binomial total" (for how many individuals we sample each time)
  \end{itemize}
  
  \endcol
  \begincol{.5\linewidth}
  ```{r distbinom, echo = FALSE}
  set.seed(42)
  hist(rbinom(n = 10000, size = 50, prob = 0.6), main = "Binomial Distribution", xlab = 
  "n = 10000, N = 50, p = 0.6")
  ```
  \endcol
\endcols

### Poisson Distribution
Another one of the **more important** distributions. It is applicable to:  

\begincols[T]
  \begincol{.5\linewidth}
  \pause
  
  * Focal objects are placed randomly in one or more dimensions  
  * A random "counting window" (usually one considering time) is placed above the sampling scheme  
  
  \pause
  It is made up of two **criteria**:  
  \pause
  
  \begin{itemize}
  \item $\lambda$ - the mean (= expectation, average count, intensity) as well as
the variance (i.e., variance = mean)
  \item $n$ - sample size
  \end{itemize}

  \endcol
  \begincol{.5\linewidth}
  ```{r distpoisson, echo = FALSE}
  set.seed(42)
  hist(rpois(1000, 5), main = "Poisson Distribution", xlab = 
  "n = 10000, Lambda = 5")
  ```
  \endcol
\endcols

## Important Measures Of Distributions
### How to Measure Distributions
\begin{center}
\textit{Not all distributions are created equally.}
\end{center}
\vspace{.3cm}
\pause
Distributions can be described via \textbf{classic parameters of descriptive statistics:}
\begin{columns}[T]
  \begin{column}{.5\linewidth}
    \begin{itemize}
    \pause
    \item Arithmetic Mean
    \item Mode
    \item Median
    \item Minimum, Maximum, Range
    \item ...
    \end{itemize}
  \end{column}
  \begin{column}{.5\linewidth}
    \begin{itemize}
    \pause
    \item Variance
    \item Standard Deviation
    \item Quantile Range
    \item \textbf{Skewness}
    \item \textbf{Kurtosis}
    \item ...
    \end{itemize}
  \end{column}
\end{columns}

\pause
\vspace{.5cm}
$\rightarrow$ Most of these are dealt with in the next seminar.

### Skewness I

\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \textit{Definition:} & Describes the symmetry and relative tail length of distributions. \\
             & \\
           \hline
             & \\
           \textit{Positive skew:} & Right-hand tail is longer than the left-hand tail\\
           \textit{Skew = 0:} & Symmetric distribution\\
           \textit{Negative skew:} & Left-hand tail is longer than the right-hand tail\\
          \end{tabular}}
        \label{Des_Stats}
\end{table}


### Skewness II
\vspace{.6cm}
\begin{columns}[T]
  \begin{column}{0.3\linewidth}
    \begin{center}
    \includegraphics[width=1\linewidth]{Figures/PosSkew.png}  
    \end{center}
    \centering\tiny{\textit{Positive Skew}}
  \end{column}
  \begin{column}{.4\linewidth}
    \begin{center}
    \includegraphics[width=1\linewidth]{Figures/Symm.png}  
    \end{center}
    \centering\tiny{\textit{Symmetric Distribution}}
  \end{column}
  \begin{column}{0.3\linewidth}
    \begin{center}
    \includegraphics[width=1\linewidth]{Figures/NegSkew.png}  
    \end{center}
    \centering\tiny{\textit{Negative Skew}}
  \end{column}
\end{columns}


### Kurtosis I
\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \textit{Definition:} & Describes the evenness/"tailedness" of distributions. \\
             & \\
           \hline
             & \\
           \textit{Positive kurtosis:} & Short-tailed distribution aka. \textit{leptokurtic}\\
           \textit{Kurtosis = 0:} & Base representation of a given distribution aka. \textit{mesokurtic}\\
           \textit{Negative kurtosis:} & Long-tailed distribution aka. \textit{platykurtic}\\
          \end{tabular}}
        \label{Des_Stats}
\end{table}

### Kurtosis II
\vspace{.6cm}
\begin{center}
\includegraphics[width=0.8\linewidth]{Figures/Kurtosis.jpg}  
\end{center}

## Estimations
### Point and Range Estimations
Point and Range estimations are parameters obtained from a \textit{sample data set} and meant to represent the \textit{population data set}.
\vspace{.3cm}
\begin{itemize}
\item With a probability of 95.5\%, the following is true
\end{itemize}
\centering $\overline{x} - 2\sigma_{\overline{x}} \leq \mu \leq \overline{x} + 2\sigma_{\overline{x}}$  

\begin{itemize}
\item With a probability of 68\%, the following is true
\end{itemize}
\centering $\overline{x} - \sigma_{\overline{x}} \leq \mu \leq \overline{x} + \sigma_{\overline{x}}$  

\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \hline
            & \\
           $\overline{x}$  & Arithmetic mean of the sample \\
           $\mu$  & Arithmetic mean of the population \\
           $\sigma_{\overline{x}}$  & Standard error of $\overline{x}$ \\
          \end{tabular}}
        \label{Des_Stats}
\end{table}

### Confidence Intervals I

\begin{tcolorbox}[colback=burgundy!5,colframe=burgundy!40!black,title=]
For multiple confidence intervals (CIs) on a level of $\alpha$ (i.e. 95\%), a proportion of $\alpha$ CIs for a given population will contain the arithmetic mean of the population.
\end{tcolorbox}

\centering [$\overline{x} - t(\alpha, df)$ ; $\overline{x} + t(\alpha, df)$]

\begin{table}[c]
        \centering
          \resizebox{\textwidth}{!}{%
          \begin{tabular}{L{0.2\linewidth}L{0.8\linewidth}}
           \hline
            & \\
           $\overline{x}$  & Arithmetic mean of the sample \\
           $\sigma_{\overline{x}}$  & Standard error of $\overline{x}$ \\
           $\alpha$ & Confidence level (usually 95\%)\\
           $df$ & Degrees of freedom\\
           $t(\alpha, df)$ & $t$-value given $\alpha$ and $df$\\
          \end{tabular}}
        \label{Des_Stats}
\end{table}

### Confidence Intervals II
The \textbf{Basics of Confidence Intervals}:
\vspace{.3cm}
\begin{itemize}
\item CIs get \textbf{larger} when:
  \begin{itemize}
  \item Smaller sample sizes
  \item Bigger spread of data values
  \item Higher statistical certainty ($\alpha$)
  \end{itemize}
\end{itemize}
\vspace{.2cm}
\begin{itemize}  
\item CIs get \textbf{narrower} when:
  \begin{itemize}
  \item Bigger sample sizes
  \item Smaller spread of data values
  \item Lower statistical certainty ($\alpha$)
  \end{itemize}
\end{itemize}

# Exercise
### `R` Environment Objects
`R` environment objects (stored as `.RData`) are highly valuable objects to any `R` user because:  

* They let you save your entire working environment  
* You cannot alter them outside of `R` (aside from deleting them)

\vspace{.2cm}

How to **create** them?  

* Use the function `save.image()` (you can specify the argument `file` for a specific name of the file)  

\vspace{.2cm}

How to **load** them?  

* Use the function `load(...)` ("..." specifies the exact path to the file on your machine)

\vspace{.2cm}

**What to do**?  

* Load the `Primer.RData` file into `R`

### Variables
Answer the following questions (take a notes for each variable!) for the variables contained within `Primer.RData`:  
\pause
\vspace{0.5cm}
\begin{itemize}
\item \textbf{What variables} am I using and:
  \begin{itemize}
  \item Of what \textit{mode} are they?
  \end{itemize}
\pause
\vspace{.3cm}
\item \textbf{What scales} apply and:
  \begin{itemize}
  \item What do they \textit{imply}?
  \item Does my data \textit{fit} these scales? Use the functions barplot() and table() when applicable!
  \linebreak
  \end{itemize}
\end{itemize}

### Distributions 

Plot the distributions of the values for the following variables:  

* `Length`  
* `Reproducing` 
* `IndividualsPassingBy`  
* `Depth`  

\vspace{.3cm}

What distributions are these? Use QQPlots or the Shapiro Test to assess normality. If you stumble upon a non-normal distribution, what else could it be?