---
title: "" # needs to be empty since it is specified in Title.Rmd
author: ""
date: ""
documentclass: "book"
fontsize: 10pt
output: 
  pdf_document:
    includes:
      in_header: Style_Exercise.tex
subparagraph: true # this is needed as a workaround to get titlesec to work
geometry: "left= 2cm,right = 2cm, bottom = 1.5cm, top = 2cm"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, cache = TRUE, cache.lazy=TRUE, tidy.opts=list(width.cutoff=70),tidy=TRUE, fig.height=4, fig.width = 12)
options(digits=2, width = 90)
```

<!--- ####### PREAMBLE -----------
--------------------------------------------->
\begin{center}
\includegraphics[width=0.42\linewidth]{Figures/UiA-logo.png}
\hspace{.05\linewidth}
\includegraphics[width=0.42\linewidth]{Figures/UL-logo.jpg}
\end{center}

\\

\vspace{5.5cm}

\begin{center}
\textsc{\LARGE{
\textbf{A Primer For Statistical Tests}\\
\vspace{0.15cm}
Biostatistics for Beginners \\
\vspace{0.3cm}
- } \vspace{0.3cm}\\
\large{BASIC STATISTICS FOR BIOLOGISTS\\
}}
\end{center}

\vspace{6.5cm}

\raggedleft
\textbf{Erik Kusch}  
\textit{PhD Student}  
Aarhus University  
Department of Bioscience  
Section for Ecoinformatics \& Biodiversity  
Center for Biodiversity and Dynamics in a Changing World (BIOCHANGE)  
Ny Munkegade 116, Building 1540  
8000 Aarhus  
Denmark  
email: erik@i-solution.de  

\raggedright

\clearpage

<!--- ####### TOC, TOF, TOT, TOL ------------------------------------------------>  
\pagestyle{plain}
\pagenumbering{roman}

\begingroup
\let\clearpage\relax
\let\cleardoublepage\relax
\begin{LARGE}
 \textbf{Summary:}\\
 \vspace{0.2cm}
\end{LARGE}
These are the solutions to the exercises contained within the handout to “A Primer For Statistical Tests” which walks you through the basics of variables, their scales and distributions. Keep in mind that there is probably a myriad of other ways to reach the same conclusions as presented in these solutions.

\setcounter{secnumdepth}{3} 
\setcounter{tocdepth}{4} 
\tableofcontents
\endgroup

<!--- ####### DOCUMENT ---------------------------------------------------------->
\pagestyle{fancy}
\pagenumbering{arabic}

# Loading the `R` Environment Object

```{r RLoad}
load("Data/Primer.RData") # load data file from Data folder
```

\let\cleardoublepage\relax

# Variables
## Finding Variables
```{r RIndentify}
ls() # list all elements in working environment
```
## `Colour`
```{r VColour}
class(Colour) # mode
barplot(table(Colour)) # fitting?
```

| Question | Answer |
|-----:|:----|
| Mode? | `r class(Colour)` |
| Which scale? | Nominal |
| What's implied? | Categorical data that can't be ordered |
| Does data fit scale? | Yes |

\clearpage

## `Depth` 
```{r VDepth, fig.height=10}
class(Depth) # mode
barplot(Depth) # fitting?
```


| Question | Answer |
|-----:|:----|
| Mode? | `r class(Depth)` |
| Which scale? | Interval/Discrete |
| What's implied? | Continuous data with a non-absence point of origin |
| Does data fit scale? | Debatable (is 0 depth absence of depth?) |

\newpage

## `IndividualsPassingBy` 
```{r VIndPass, fig.height=10}
class(IndividualsPassingBy) # mode
barplot(IndividualsPassingBy) # fitting?
```

| Question | Answer |
|-----:|:----|
| Mode? | `r class(IndividualsPassingBy)` |
| Which scale? | Integer |
| What's implied? | Only integer numbers with an absence point of origin |
| Does data fit scale? | Yes |

\newpage

## `Length` 
```{r VLength, fig.height=10}
class(Length) # mode
barplot(Length) # fitting?
```

| Question | Answer |
|-----:|:----|
| Mode? | `r class(Length)` |
| Which scale? | Relation/Ratio |
| What's implied? | Continuous data with an absence point of origin |
| Does data fit scale? | Yes |

\newpage

## `Reproducing` 
```{r VRepro, fig.height=10}
class(Reproducing) # mode
barplot(Reproducing) # fitting?
```

| Question | Answer |
|-----:|:----|
| Mode? | `r class(Reproducing)` |
| Which scale? | Integer |
| What's implied? | Only integer numbers with an absence point of origin |
| Does data fit scale? | Yes |

\newpage

## `Sex`
```{r VSex, fig.height=10}
class(Sex) # mode
barplot(table(Sex)) # fitting?
```

| Question | Answer |
|-----:|:----|
| Mode? | `r class(Sex)` |
| Which scale? | Binary |
| What's implied? | Only two possible outcomes |
| Does data fit scale? | Yes |

\newpage

## `Size` 
```{r VSize, fig.height=10}
class(Size) # mode
barplot(table(Size)) # fitting?
```

| Question | Answer |
|-----:|:----|
| Mode? | `r class(Size)` |
| Which scale? | Ordinal |
| What's implied? | Categorical data that can be ordered |
| Does data fit scale? | Yes |

\newpage

## `Temperature` 
```{r VTemp, fig.height=10}
class(Temperature) # mode
barplot(Temperature) # fitting?
```

| Question | Answer |
|-----:|:----|
| Mode? | `r class(Temperature)` |
| Which scale? | Interval/Discrete |
| What's implied? | Continuous data with a non-absence point of origin |
| Does data fit scale? | Yes (the data is clearly recorded in degree Celsius) |

\newpage

# Distributions

## `Length`  
```{r DLength, fig.height=10}
plot(density(Length)) # distribution plot
shapiro.test(Length) # normality check
```
The data is **normal distributed**.

\newpage

## `Reproducing` 
```{r DRepro, fig.height=10}
plot(density(Reproducing)) # distribution
shapiro.test(Reproducing) # normality check
```
The data is **binomial distributed** (i.e. "How many individuals manage to reproduce") but looks **normal distributed**. The normal distribution doesn't make sense here because it implies continuity whilst the data only comes in integers.

\newpage

## `IndividualsPassingBy`  
```{r DIndiv, fig.height=10}
plot(density(IndividualsPassingBy)) # distribution
shapiro.test(IndividualsPassingBy) # normality check
```
The data is **poisson distributed** (i.e. "How many individuals pass by an observer in a given time frame?").

\newpage

## `Depth` 
```{r DDepth, fig.height=10}
plot(density(Depth)) # distribution
```
The data is **uniform distributed**. You don't know this distribution class from the lectures and I only wanted to confuse you with this to show you that there's much more out there than I can show in our lectures.